---
title: "Benchmark segmentation across imaging ST platforms"
output: html_document
date: "2026-02-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/SpatialBench_segmentation_comparison/Data")
```

# Load packages

```{r}
suppressPackageStartupMessages({
  library(Seurat) # v5.4.0
  library(future)
  library(future.apply)
  library(ggplot2)
  library(ggrepel)
  library(scales) # for improving plot visualisation
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(purrr)
  library(data.table)
  library(patchwork) # for plot aggregation
})
```

# Load data

## Load Seurat objects

```{r}
main_dir <- "/vast/scratch/users/zhang.ji/data"
merscope_sub_dir <- "Merscope_seurat/vizgen_ctrl_wt_ko_annotated.rds"
merscope_vizgen <- readRDS(file.path(main_dir, "default_segmentation", merscope_sub_dir))
merscope_proseg <- readRDS(file.path(main_dir, "proseg_segmentation", merscope_sub_dir))
merscope_cellpose <- readRDS(file.path(main_dir, "cellpose_segmentation", merscope_sub_dir))
```

```{r}
xenium_uni_sub_dir <- "Xenium_seurat/unimodal/xenium_ctrl_wt_ko_annotated.rds"
xenium_uni_default <- readRDS(file.path(main_dir, "default_segmentation", xenium_uni_sub_dir))
xenium_uni_proseg <- readRDS(file.path(main_dir, "proseg_segmentation", xenium_uni_sub_dir))
```

```{r}
xenium_multi_sub_dir <- c(
  "default_segmentation/Xenium_seurat/multimodal/xenium_ctrl_wt_ko_annotated.rds",
  "proseg_segmentation/Xenium_seurat/multimodal/xenium_ctrl_wt_ko_annotated.rds",
  "cellpose_segmentation/Xenium_seurat/xenium_ctrl_wt_ko_annotated.rds"
)
xenium_multi_default <- readRDS(file.path(main_dir, xenium_multi_sub_dir[1]))
xenium_multi_proseg <- readRDS(file.path(main_dir, xenium_multi_sub_dir[2]))
xenium_multi_cellpose <- readRDS(file.path(main_dir, xenium_multi_sub_dir[3]))
```

## Load sample-level and cell-level data information

```{r}
sample_df <- read.csv("sample_df.csv") %>% 
  mutate(
    platform_version = case_when(
      is.na(model) ~ platform, 
      .default = paste(platform, model)), 
    segmentation_clean = case_when(
      segmentation %in% c("Cellpose1", "Nuclear expansion") ~ "Default",
      .default = segmentation
    )
  )
sample_df$segmentation_clean <- factor(
  sample_df$segmentation_clean,
  levels = c("Default", "Proseg", "Cellpose2")
)

assignment_info <- read.csv("assignment_info.csv") %>% 
  mutate(
    platform_version = case_when(
      is.na(model) ~ platform, 
      .default = paste(platform, model)), 
    segmentation_clean = case_when(
      segmentation %in% c("Cellpose1", "Nuclear expansion") ~ "Default",
      .default = segmentation
    )
  )
assignment_info$segmentation_clean <- factor(
  assignment_info$segmentation_clean,
  levels = c("Default", "Proseg", "Cellpose2")
)

cell_df <- data.table::fread("cell_df.csv.gz") %>% 
  mutate(
    platform_version = case_when(
      model == "" ~ platform, 
      .default = paste(platform, model)), 
    segmentation_clean = case_when(
      segmentation %in% c("Cellpose1", "Nuclear expansion") ~ "Default",
      .default = segmentation
    )
  )
cell_df$segmentation_clean <- factor(
  cell_df$segmentation_clean,
  levels = c("Default", "Proseg", "Cellpose2")
)
```

# Load functions

```{r}
source("../Scripts/functions/benchmarking_functions.r")
```

# Basic statistics

## Sample-level quality metrics

Subset to common samples and segmentation methods across platforms.

```{r}
# sample-level data frame
common_samples <- sample_df %>% 
  filter(platform_version != "Xenium unimodal") %>% 
  pull(sample_id) %>% 
  unique()

sample_df_filtered <- sample_df %>% 
  filter(
    platform_version != "Xenium unimodal", 
    sample_id %in% common_samples
  )
head(sample_df_filtered)
```

```{r}
# assignment information data
common_samples <- assignment_info %>% 
  filter(platform_version == "Xenium multimodal") %>% 
  pull(sample) %>% 
  unique()

assignment_info_filtered <- assignment_info %>% 
  filter(
    platform_version != "Xenium unimodal",
    sample %in% common_samples,
  )
head(assignment_info)
```


### Total cell counts

```{r}
total_cell_counts_filtered <- sample_boxplot(
  sample_df_filtered, 
  metric = "cell_count", 
  label = "Total cell counts"
)
total_cell_counts_filtered
```

### Total transcriopt counts

```{r}
total_transcript_counts_filtered_common <- sample_boxplot(
  sample_df_filtered, 
  metric = "common_gene_counts", 
  label = "Total transcript counts (common genes)"
)
total_transcript_counts_filtered_common
```

### Percentage of transcripts assigned

```{r}
pt_assignment_filtered_common <- sample_boxplot(
  assignment_info_filtered,
  metric = "pt_assignment_common", 
  label = "% Transcripts assigned to cell (common genes)",
  jitter_width = 0.001,
  jitter_height = 0.001
) + coord_cartesian(ylim = c(0, 100))
pt_assignment_filtered_common
```

### All-in-one

```{r fig.width=12, fig.height=4}
sample_level_quality_metrics_common_filtered <- total_cell_counts_filtered | total_transcript_counts_filtered_common | pt_assignment_filtered_common 
ggsave("~/data/figures/sample_level_quality_metrics_common_filtered.pdf", plot = sample_level_quality_metrics_common_filtered, width = 12, height = 4)

sample_level_quality_metrics_common_filtered
```

## Cell-level quality metrics

Subset to common samples and segmentation across platforms.

```{r}
common_samples <- cell_df %>% 
  filter(platform_version == "Xenium multimodal") %>% 
  pull(sample) %>% 
  unique()

cell_df_filtered <- cell_df %>% 
  filter(
    platform_version != "Xenium unimodal",
    sample %in% common_samples
  )

head(cell_df_filtered)
```


### Counts per cell

Inspect extreme high counts per cell across samples

```{r}
cell_df_filtered %>% 
  group_by(platform_version, segmentation_clean, sample) %>% 
  summarise(pt = mean(nCounts > 3000) * 100,
            .groups = "drop") %>% 
  arrange(desc(pt))
```


```{r fig.width=12, fig.height=4}
counts_per_cell_filtered_common <- cell_violin_plot(
  cell_df_filtered %>% 
    filter(common_gene_counts <= 3000),
  metric = "common_gene_counts",
  label = "Counts per cell (common genes)"
)
counts_per_cell_filtered_common
```

```{r}
median_counts_per_cell_filtered_common <- cell_median_boxplot(
  cell_df_filtered,
  metric = "common_gene_counts",
  label = "counts per cell (common genes)"
)
median_counts_per_cell_filtered_common
```

### Genes per cell

```{r fig.width=12, fig.height=4}
genes_per_cell_filtered_common <- cell_violin_plot(
  cell_df_filtered,
  metric = "common_feature_counts",
  label = "Genes per cell (common genes)"
)
genes_per_cell_filtered_common
```

```{r}
median_genes_per_cell_filtered_common <- cell_median_boxplot(
  cell_df_filtered,
  metric = "common_feature_counts",
  label = "genes per cell (common genes)"
) + scale_y_continuous(limits = c(0, 91), n.breaks = 8) + 
  geom_hline(yintercept = 18, linetype = "dashed", colour = "brown")
median_genes_per_cell_filtered_common
```

### All-in-one

```{r fig.width=15, fig.height=8}
layout <- "
AAAAABB
#######
CCCCCDD
"

cell_level_quality_metrics_filtered_common <- wrap_plots(list(
  counts_per_cell_filtered_common, 
  median_counts_per_cell_filtered_common,
  genes_per_cell_filtered_common,
  median_genes_per_cell_filtered_common
)) + 
  plot_layout(design = layout, heights = c(1, 0.2, 1))

ggsave("~/data/figures/cell_level_quality_metrics_filtered_common.pdf", plot = cell_level_quality_metrics_filtered_common, width = 15, height = 8)
cell_level_quality_metrics_filtered_common
```

## Morphological metrics

### Cell area

Inspect cells with extreme areas based on identified cell types (they should not be greater than 500 $\mu m^2$).

```{r}
cell_df_filtered %>% 
  group_by(platform_version, segmentation, sample) %>% 
  summarise(pt = mean(cell_area > 500) * 100,
            .groups = "drop") %>% 
  arrange(desc(pt))
```

```{r fig.width=12, fig.height=4}
cell_area_filtered <- cell_violin_plot(
  cell_df_filtered %>% 
    filter(cell_area <= 500),
  metric = "cell_area",
  label = ""
) + ylab(expression("Cell area ("*µm^2*")"))
cell_area_filtered
```

```{r}
median_cell_area_filtered <- cell_median_boxplot(
  cell_df_filtered,
  metric = "cell_area",
  label = ""
) + ylab(expression("Median cell area ("*µm^2*")"))
median_cell_area_filtered
```

#### Cell-type specific

```{r}
median_celltype_filtered_df <- cell_df_filtered %>% 
  group_by(platform_version, segmentation_clean, sample, cell_type) %>% 
  summarise(
    area_median = median(cell_area), 
    nCounts_median = median(common_gene_counts),
    nFeatures_median = median(common_feature_counts),
    .groups = "drop"
  )
```

```{r fig.width=20, fig.height=4}
median_celltype_area_filtered_plot <- ggplot(
  median_celltype_filtered_df, 
  aes(x = platform_version, y = area_median, fill = segmentation_clean)) +
  geom_boxplot(outliers = F) +
  geom_jitter(
    position = position_jitterdodge(
      jitter.width = 0.1, 
      jitter.height = 0.1, 
      dodge.width = 0.75
    ), 
    size = 0.5
  ) + 
  scale_fill_manual(values = colour_panel) + 
  facet_wrap(~cell_type, nrow = 1) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +  # y starts at 0
  coord_cartesian(ylim = c(0, NA)) +  # force y=0 lower bound, free upper bound
  labs(
    x = "Platform", 
    y = expression("Median cell area ("*µm^2*")")
  ) + 
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
ggsave("~/data/figures/median_celltype_area_filtered_plot.pdf", plot = median_celltype_area_filtered_plot, width = 20, height = 4)

median_celltype_area_filtered_plot
```

Add cell-type specific reference area ranges.

```{r fig.width=20, fig.height=4}
# set cell order to display
median_celltype_filtered_df$cell_type <- factor(
  median_celltype_filtered_df$cell_type,
  levels = c("B cells", "T cells", "GC B cells", "Plasma cells",
             "Macrophages", "Neutrophils", "Erythrocyte-like", 
             "Dendritic cells", "Endothelial cells", "NK cells",
             "Fibroblastic reticular cells")
)

# with referenced cell area
ref_cell_area <- data.frame(
  cell_type = c("T cells", "B cells", "Macrophages", "Erythrocyte-like", 
                "Neutrophils", "GC B cells", "Plasma cells", "Dendritic cells", 
                "Endothelial cells", "NK cells", "Fibroblastic reticular cells"),
  low_bound = c(37.83, 15.9, 103.87, 29.71, 126.68, NA, 38.48, rep(NA, 4)),
  up_bound = c(75.12, 75.12, 203.58, 35.78, 169.72, NA, 201.06, rep(NA, 4))
)
ref_cell_area$cell_type <- factor(
  ref_cell_area$cell_type,
  levels = c("B cells", "T cells", "GC B cells", "Plasma cells",
             "Macrophages", "Neutrophils", "Erythrocyte-like", 
             "Dendritic cells", "Endothelial cells", "NK cells",
             "Fibroblastic reticular cells")
)

median_celltype_area_filtered_with_ref <- ggplot(
  median_celltype_filtered_df, 
  aes(x = platform_version, y = area_median, fill = segmentation_clean)
) +
  geom_boxplot(outliers = F) +
  geom_jitter(
    position = position_jitterdodge(
      jitter.width = 0.1,    # horizontal spread of dots
      jitter.height = 0.1,
      dodge.width = 0.75     # must match boxplot's dodge width (default is 0.75)
    ),
    size = 0.5
  ) + 
  scale_fill_manual(values = colour_panel) + 
  geom_hline(data = ref_cell_area, 
             aes(yintercept = low_bound),
             linetype = "dashed", 
             color = "brown") + 
  geom_hline(data = ref_cell_area, 
           aes(yintercept = up_bound),
           linetype = "dashed", 
           color = "brown") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), n.breaks = 5) +  # y starts at 0
  coord_cartesian(ylim = c(0, NA)) +
  # facet_grid(rows = vars(platform_version), cols = vars(cell_type), scales = "fixed") + 
  facet_wrap(~cell_type, nrow = 1) + 
  labs(x = "Platform", y = expression("Median cell area ("*µm^2*")")) + 
  theme_bw() +
  theme(
    legend.position = "none", 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("~/data/figures/median_celltype_area_filtered_with_ref.pdf", plot = median_celltype_area_filtered_with_ref, width = 20, height = 4)
median_celltype_area_filtered_with_ref
```

Violin plot across samples

```{r fig.width=25, fig.height=6}
ggplot(
    cell_df_filtered %>% filter(cell_area <= 500), 
    aes(x = sample, y = cell_area, fill = segmentation_clean)
  ) + 
    geom_violin(trim = FALSE, scale = "width") + 
    geom_boxplot(aes(group = interaction(sample, segmentation_clean)), 
                 width = 0.15, outlier.shape = NA, fill = "white", color = "black",
                 position = position_dodge(0.9)) +
    scale_fill_manual(values = colour_panel) +
    facet_grid(rows = vars(platform_version), cols = vars(cell_type), scales = "fixed") + 
    labs(x = "Samples",
         y = expression("Cell area ("*µm^2*")"),
         fill = "segmentation_clean") + 
    scale_y_continuous(n.breaks = 8) + 
    theme_bw() + 
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )
```


Selected cells: B cells, T cells, Macrophages, Neutrophils

```{r}
median_celltype_sub_df <- median_celltype_filtered_df %>% 
  filter(cell_type %in% c("B cells", "T cells", "Macrophages", "Neutrophils"))
ref_cell_area_sub <- ref_cell_area %>% 
  filter(cell_type %in% c("B cells", "T cells", "Macrophages", "Neutrophils"))

median_celltype_area_sub_ref <- ggplot(
  median_celltype_sub_df, 
  aes(x = segmentation_clean, y = area_median, fill = segmentation_clean)
) +
  geom_boxplot(outliers = F) +
  geom_jitter(width = 0.1, height = 0.1) + 
  scale_fill_manual(values = colour_panel) + 
  geom_hline(data = ref_cell_area_sub, 
             aes(yintercept = low_bound),
             linetype = "dashed", 
             color = "brown") + 
  geom_hline(data = ref_cell_area_sub, 
           aes(yintercept = up_bound),
           linetype = "dashed", 
           color = "brown") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), n.breaks = 5) +  # y starts at 0
  coord_cartesian(ylim = c(0, NA)) +
  facet_grid(rows = vars(platform_version), cols = vars(cell_type), scales = "fixed") + 
    # force y=0 lower bound, free upper bound
  labs(x = "Segmentation", y = expression("Median cell area ("*µm^2*")")) + 
  theme_bw() +
  theme(legend.position = "none")

median_celltype_area_sub_ref
```


Statistical testing on differences in cell area across segmentation

```{r}
library(broom)
models_df <- median_celltype_filtered_df %>% 
  group_by(platform_version, cell_type) %>% 
  tidyr::nest() %>% 
  mutate(model = map(data, ~aov(area_median ~ segmentation_clean, data = .x)))

sig_celltypes_anova <- models_df %>% 
  mutate(anova = map(model, ~ tidy(anova(.x)))) %>% 
  unnest(anova) %>% 
  filter(term == "segmentation_clean", p.value < 0.05) %>% 
  select(platform_version, cell_type)

sig_celltypes_anova_list <- sig_celltypes_anova %>% 
  distinct() %>% 
  group_by(platform_version) %>%
  summarise(cell_types = list(cell_type), .groups = "drop") %>%
  deframe()

emm_results <- models_df %>%
  mutate(
    emm = map(model, ~ emmeans(.x, ~ segmentation_clean)),
    tukey = map(emm, ~ pairs(.x, adjust = "tukey")),
    tukey_sig = map(
      tukey,
      ~ {
        s <- summary(.x)
        s[s$p.value < 0.05, ]
      }
    )
  )

sig_contrast_df <- emm_results %>%
  select(platform_version, cell_type, tukey_sig) %>% 
  unnest(tukey_sig)

sig_contrast_list <- sig_contrast_df %>%
  group_by(platform_version, cell_type) %>%
  summarise(contrasts = list(contrast), .groups = "drop") %>%
  group_by(platform_version) %>%
  summarise(cell_contrasts = list(setNames(contrasts, cell_type)), .groups = "drop") %>%
  deframe() 

sig_contrast_list
```


```{r fig.width=20, fig.height=4}
median_celltype_counts_filtered_plot <- ggplot(
  median_celltype_filtered_df, 
  aes(x = platform_version, y = nCounts_median, fill = segmentation_clean)) +
  geom_boxplot(outliers = F) +
  geom_jitter(
    position = position_jitterdodge(
      jitter.width = 0.1, 
      jitter.height = 0.1,
      dodge.width = 0.75
    ), 
    size = 0.5
  ) + 
  scale_fill_manual(values = colour_panel) + 
  facet_wrap(~cell_type, nrow = 1) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +  # y starts at 0
  coord_cartesian(ylim = c(0, NA)) +  # force y=0 lower bound, free upper bound
  labs(
    x = "Platform", 
    y = "Median counts per cell (common genes)"
  ) + 
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
ggsave("~/data/figures/median_celltype_counts_filtered_plot.pdf", plot = median_celltype_counts_filtered_plot, width = 20, height = 4)

median_celltype_counts_filtered_plot
```

```{r fig.width=20, fig.height=4}
median_celltype_features_filtered_plot <- ggplot(
  median_celltype_filtered_df, 
  aes(x = platform_version, y = nFeatures_median, fill = segmentation_clean)) +
  geom_boxplot(outliers = F) +
  geom_jitter(
    position = position_jitterdodge(
      jitter.width = 0.1, 
      jitter.height = 0.1,
      dodge.width = 0.75
    ), 
    size = 0.5
  ) + 
  scale_fill_manual(values = colour_panel) + 
  facet_wrap(~cell_type, nrow = 1) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), n.breaks = 6) +  # y starts at 0
  coord_cartesian(ylim = c(0, 91)) +
  labs(
    x = "Platform", 
    y = "Median genes per cell (common genes)"
  ) + 
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
ggsave("~/data/figures/median_celltype_features_filtered_plot.pdf", plot = median_celltype_features_filtered_plot, width = 20, height = 4)

median_celltype_features_filtered_plot
```


#### Heatmap and statistical testing

```{r}
suppressPackageStartupMessages({
  library(boot)
  library(coin)
  library(ComplexHeatmap)
  library(circlize)
  library(grid)
  library(combinat) # for combn()
})

# exact permutation test on log2 median ratio
perm_test_median_ratio <- function(x, y, log2_scale = TRUE) {
  n_x <- length(x)
  n_y <- length(y)
  pooled <- c(x, y)
  n_pooled <- length(pooled)
  
  # observed statistic
  obs_stat <- ifelse(log2_scale, log2(median(x) / median(y)), median(x) / median(y))
  
  # all ways to choose indices for x
  perm_indices <- combinat::combn(n_pooled, n_x)   # n_x x C(n_pooled, n_x) matrix
  
  # compute statistic for every split
  perm_stats <- apply(perm_indices, 2, function(idx) {
    perm_x <- pooled[idx]
    perm_y <- pooled[-idx]
    log2(median(perm_x) / median(perm_y))
  })
  
  # two-sided p-value: proportion of permuted stats as or more extreme
  p_val <- mean(abs(perm_stats) >= abs(obs_stat))
  
  return(list(
    obs_stat = obs_stat,
    p_val = p_val
  ))
}

# function to compute scaled effect, p-value, CI across platforms
analyse_metric_celltype_across_platform <- function(subset_list, log2_scale = TRUE) {
  # values per platform
  xen_vals <- subset_list$xen_vals
  mer_vals <- subset_list$mer_vals
  
  if (length(mer_vals) == 0 | length(xen_vals) == 0) {
    return(list(scaled_effect = NA, p_val = NA))
  }
  
  # perform permutation test on log2 median ratio between MERSCOPE and Xenium multimodal platforms
  res <- perm_test_median_ratio(x = xen_vals, y = mer_vals, log2_scale = log2_scale)
  
  return(list(
    metric = subset_list$metric,
    col_label = subset_list$col_label,
    scaled_effect = res$obs_stat,
    p_val = res$p_val
  ))
}

# function to compute scaled effect, p-value, CI across segmentation
analyse_metric_celltype_across_seg <- function(subset_list, log2_scale = TRUE) {
  # values per segmentation
  default_vals <- subset_list$default_vals
  proseg_vals <- subset_list$proseg_vals
  cellpose_vals <- subset_list$cellpose_vals
  
  if (any(length(default_vals) == 0, length(proseg_vals) == 0, length(cellpose_vals)) == 0) {
    return(list(scaled_effect = NA, p_val = NA))
  }
  
  # perform permutation test on log2 median ratio between (Default, Cellpose2) and (Proseg, Cellpose2)
  res_default <- perm_test_median_ratio(x = default_vals, y = cellpose_vals, log2_scale = log2_scale)
  res_proseg <- perm_test_median_ratio(x = proseg_vals, y = cellpose_vals, log2_scale = log2_scale)
  
  return(list(
    metric = subset_list$metric,
    col_label = subset_list$col_label,
    scaled_effect_default = res_default$obs_stat,
    scaled_effect_proseg = res_proseg$obs_stat,
    p_val_default = res_default$p_val,
    p_val_proseg = res_proseg$p_val
  ))
}

plot_area_heatmap_across_plat <- function(df, log2_scale = TRUE) {
  # prepare plot data
  plot_df <- df %>% 
    group_by(platform_version, segmentation_clean, cell_type, sample) %>% 
    summarise(
      "Cell area" = median(cell_area),
      "Number of transcripts" = median(common_gene_counts),
      "Number of genes" = median(common_feature_counts),
      .groups = "drop"
    ) %>% 
    ungroup()
  
  # list of platforms
  platforms <- unique(plot_df$platform_version)
  if (length(platforms) != 2) {
    stop("This function assums exactly 2 platforms")
  }
  if (!all(platforms %in% c("MERSCOPE", "Xenium multimodal"))) {
    stop("The platforms must be MERSCOPE and Xenium multimodal")
  }
  
  # define the column order of heatmap
  cell_types <- levels(plot_df$cell_type)
  segmentations <- c("Default", "Proseg", "Cellpose2")
  col_order <- data.frame(
    cell_type = rep(cell_types, each = length(segmentations)),
    segmentation = rep(segmentations, times = length(cell_types))
  ) %>% 
    mutate(col_label = paste(cell_type, segmentation, sep = "|"))
  
  # pre-split data for each metric and cell type
  # nested list: split_df[[metric & cell type]]
  metrics <- c("Cell area", "Number of transcripts", "Number of genes")
  split_df <- plot_df %>% 
    pivot_longer(cols = metrics, names_to = "metric", values_to = "value") %>% 
    group_by(segmentation_clean, metric, cell_type) %>% 
    group_split() %>% 
    lapply(function(df) {
      list(
        metric = unique(df$metric),
        col_label = paste(unique(df$cell_type), unique(df$segmentation_clean), sep = "|"),
        mer_vals = df %>% filter(platform_version == "MERSCOPE") %>% pull(value),
        xen_vals = df %>% filter(platform_version == "Xenium multimodal") %>% pull(value)
      )
    })
  
  # apply permutation test to all subsets
  res_list <- lapply(split_df, function(df) analyse_metric_celltype_across_platform(df, log2_scale = log2_scale))
  
  # Add results to matrices for heatmap
  heatmap_values <- matrix(
    NA, nrow = length(metrics), ncol = length(col_order$col_label), 
    dimnames = list(metrics, col_order$col_label)
  )
  pvals <- matrix(
    NA, nrow = length(metrics), ncol = length(col_order$col_label), 
    dimnames = list(metrics, col_order$col_label)
  )
  for (res in res_list) {
    r <- res$metric
    c <- res$col_label
    heatmap_values[r, c] <- res$scaled_effect
    pvals[r, c] <- res$p_val
  }
  
  # multiple testing correction (BH)
  pvals_adj <- pvals
  # metric-wise adjustment
  for (i in seq_len(nrow(pvals))) {
    for (seg in segmentations) {
      col <- grepl(seg, colnames(pvals_adj))
      pvals_adj[i, col] <- p.adjust(pvals[i, col], method = "BH")
    }
  }
  
  # mark significance
  sig_mat <- ifelse(pvals_adj < 0.05, TRUE, FALSE)
  
  segmentation_colours <- c(
    "Default" = "#F8766D",
    "Proseg" = "#00BA38",
    "Cellpose2" = "#619CFF"
  )
  
  # set annotation for segmentation colour bar
  seg_bar <- HeatmapAnnotation(
    Segmentation = col_order$segmentation,
    col = list(Segmentation = segmentation_colours),
    show_annotation_name = TRUE,
    annotation_name_side = "right",
    annotation_name_gp = gpar(fontsize = 10),
    simple_anno_size = unit(5, "mm"),
    annotation_legend_param = list(
      Segmentation = list(
        title = "Segmentation",
        at = segmentations,
        labels = segmentations,
        nrow = 1,
        legend_gp = gpar(fill = segmentation_colours)
      )
    )
  )

  col_fun <- colorRamp2(
    c(0, 4, 8),
    c("white", "#FDD18E", "#FDA203")
  )
  
  # column split by cell pair
  col_split <- factor(col_order$cell_type, levels = cell_types)
  
  # draw heatmap
  ht <- Heatmap(
    heatmap_values,
    name = "Log2 median ratio",
    col = col_fun,
    
    # column settings
    column_split = col_split,
    column_gap = unit(1.5, "mm"),
    column_title_gp   = gpar(fontsize = 10),
    column_title_side = "top",            # segmentation labels at top
    show_column_names = FALSE,            # hide individual segmentation names
    column_title_rot = 45,
    cluster_columns   = FALSE,
    cluster_column_slices = FALSE,
    
    # row settings
    cluster_rows      = FALSE,
    row_names_side    = "right",
    row_names_gp      = gpar(fontsize = 10),
    
    # cell size
    width = unit(ncol(heatmap_values) * 5, "mm"),
    height = unit(nrow(heatmap_values) * 5, "mm"),
    
    # top annotation (segmentation bar below cell pair title)
    top_annotation = seg_bar,
    
    # significance stars
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (sig_mat[i, j]) {
        grid.text("*", x, y, gp = gpar(fontsize = 10, col = "black"))
      }
    },
    
    # borders
    border            = TRUE,
    rect_gp           = gpar(col = "white", lwd = 0.5),
    
    # legend
    heatmap_legend_param = list(
      col_fun = col_fun,
      title = "Median ratio",
      at = seq(0, 8, length.out = 5),
      direction = "horizontal",
      title_position = "topcenter",
      legend_width  = unit(5, "cm"),
      labels_gp     = gpar(fontsize = 9)
    )
  ) 
  
  draw(ht,
       heatmap_legend_side    = "bottom",
       annotation_legend_side = "bottom",
       merge_legend           = TRUE)
}


plot_area_heatmap_across_seg <- function(df, log2_scale = TRUE) {
  # prepare plot data
  plot_df <- df %>% 
    group_by(platform_version, segmentation_clean, cell_type, sample) %>% 
    summarise(
      "Cell area" = median(cell_area),
      "Number of transcripts" = median(common_gene_counts),
      "Number of genes" = median(common_feature_counts),
      .groups = "drop"
    ) %>% 
    ungroup()
  
  # list of segmentation
  segmentations <- levels(plot_df$segmentation_clean)
  if (length(segmentations) != 3) {
    stop("This function assumes exactly 3 segmentation methods")
  }
  if (!all(segmentations %in% c("Default", "Proseg", "Cellpose2"))) {
    stop("The segmentation methods must be Default, Proseg and Cellpose2")
  }
  
  # define the column order of heatmap
  platforms <- unique(plot_df$platform_version)
  cell_types <- levels(plot_df$cell_type)
  col_order <- data.frame(
    cell_type = rep(cell_types, each = length(platforms)),
    platform = rep(platforms, times = length(cell_types))
  ) %>% 
    mutate(col_label = paste(cell_type, platform, sep = "|"))
  
  # pre-split data for each metric and cell type
  # nested list: split_df[[metric & cell type]]
  metrics <- c("Cell area", "Number of transcripts", "Number of genes")
  split_df <- plot_df %>% 
    pivot_longer(cols = metrics, names_to = "metric", values_to = "value") %>% 
    group_by(platform_version, metric, cell_type) %>% 
    group_split() %>% 
    lapply(function(df) {
      list(
        metric = unique(df$metric),
        col_label = paste(unique(df$cell_type), unique(df$platform_version), sep = "|"),
        default_vals = df %>% filter(segmentation_clean == "Default") %>% pull(value),
        proseg_vals = df %>% filter(segmentation_clean == "Proseg") %>% pull(value),
        cellpose_vals = df %>% filter(segmentation_clean == "Cellpose2") %>% pull(value)
      )
    })
  
  # apply permutation test to all subsets
  res_list <- lapply(split_df, function(df) analyse_metric_celltype_across_seg(df, log2_scale = log2_scale))
  
  # Add results to matrices for heatmap
  row_label <- paste(rep(metrics, each = 2), rep(c("Default", "Proseg"), times = 2), sep = "|")
  heatmap_values <- matrix(
    NA, nrow = length(row_label), ncol = length(col_order$col_label), 
    dimnames = list(row_label, col_order$col_label)
  )
  pvals <- matrix(
    NA, nrow = length(row_label), ncol = length(col_order$col_label), 
    dimnames = list(row_label, col_order$col_label)
  )
  for (res in res_list) {
    r1 <- paste(res$metric, "Default", sep = "|")
    r2 <- paste(res$metric, "Proseg", sep = "|")
    c <- res$col_label
    heatmap_values[r1, c] <- res$scaled_effect_default
    heatmap_values[r2, c] <- res$scaled_effect_proseg
    pvals[r1, c] <- res$p_val_default
    pvals[r2, c] <- res$p_val_proseg
  }
  
  # multiple testing correction (BH)
  pvals_adj <- pvals
  # metric-wise adjustment
  for (i in seq_len(nrow(pvals))) {
    for (platform in platforms) {
      col <- grepl(platform, colnames(pvals_adj))
      pvals_adj[i, col] <- p.adjust(pvals[i, col], method = "BH")
    }
  }
  
  # mark significance
  sig_mat <- ifelse(pvals_adj < 0.05, TRUE, FALSE)
  
  # colour scale
  col_fun <- colorRamp2(
    c(-1, 0, 4),
    c("#2166AC", "white", "#FDA203")   
  )
  
  # row split by comparison
  row_split <- factor(
    rep(metrics, each = 2),
    levels = metrics
  )
  
  # row annotation (coloured by comparison)
  # each group of 2 rows, row1 = Default vs Cellpose2, row2 = Proseg vs Cellpose2
  comparison_colours <- rep(c("#F8766D", "#00BA38"), length(metrics))
  row_anno <- rowAnnotation(
    Comparison = anno_simple(
      comparison_colours,
      col = c("#F8766D" = "#F8766D", "#00BA38" = "#00BA38"),
      width = unit(5, "mm")
    ),
    show_annotation_name = FALSE
  )
  
  # column split by cell types
  col_split <- factor(col_order$cell_type, levels = cell_types)
  
  # column annotation (coloured by platform)
  # each cell type has 2 sub-columns (one per platform)
  platform_colours <- rep(c("#764698", "#363a64"), length(cell_types))
  top_anno <- HeatmapAnnotation(
    Platform = anno_simple(
      platform_colours,
      col = c("#764698" = "#764698", "#363a64" = "#363a64"),
      height = unit(5, "mm")
    ),
    show_annotation_name = FALSE
  )
  
  # cell function: draw star if significant
  cell_fun <- function(j, i, x, y, width, height, fill) {
    if (sig_mat[i, j]) {
      grid.text("*", x, y, gp = gpar(fontsize = 8, col = "black"))
    }
  }
  
  # draw heatmap
  ht <- Heatmap(
    heatmap_values,
    name = "Log2 median ratio",
    col  = col_fun,
    
    # column settings
    cluster_columns   = FALSE,
    column_split      = col_split,
    column_gap        = unit(1.5, "mm"),
    column_title_gp   = gpar(fontsize = 10),
    column_title_side = "top",
    column_title_rot  = 45,
    show_column_names = FALSE,
    
    # row settings
    cluster_rows      = FALSE,
    show_row_names    = FALSE,          # hide pipe-separated row labels
    row_split         = row_split,
    row_title_side    = "right",        # metric names appear on the right
    row_title_rot     = 0,
    row_title_gp      = gpar(fontsize = 10),
    row_gap           = unit(2, "mm"),
    
    # cell size
    width  = unit(ncol(heatmap_values) * 5, "mm"),
    height = unit(nrow(heatmap_values) * 5, "mm"),
    
    # annotations
    top_annotation   = top_anno,
    right_annotation = row_anno,
    
    # significance stars
    cell_fun = cell_fun,
    
    # borders
    border  = TRUE,
    rect_gp = gpar(col = "white", lwd = 0.5),
    
    # legend
    heatmap_legend_param = list(
      title          = "Median ratio",
      at             = c(-1, 0, 1, 2, 3, 4),
      direction      = "horizontal",
      title_position = "topcenter",
      legend_width   = unit(5, "cm"),
      labels_gp      = gpar(fontsize = 9)
    )
  )
  
  # Platform legend
  lgd_platform <- Legend(
    title      = "Platform",
    labels     = c("MERSCOPE", "Xenium mutimodal"),
    legend_gp  = gpar(fill = c("#764698", "#363a64")),
    title_gp   = gpar(fontsize = 10, fontface = "bold"),
    labels_gp  = gpar(fontsize = 9),
    direction  = "horizontal",
    nrow       = 1
  )
  
  # Segmentation comparison legend
  lgd_comparison <- Legend(
    title      = "Segmentation comparison",
    labels     = c("Default vs Cellpose2", "Proseg vs Cellpose2"),
    legend_gp  = gpar(fill = c("#F8766D", "#00BA38")),
    title_gp   = gpar(fontsize = 10, fontface = "bold"),
    labels_gp  = gpar(fontsize = 9),
    direction  = "horizontal",
    nrow       = 1
  )
  
  # Pack legends side by side at the bottom
  pd <- packLegend(
    lgd_platform,
    lgd_comparison,
    direction = "horizontal",
    gap       = unit(1, "cm")
  )
  
  grid.grabExpr({
    draw(ht, heatmap_legend_side = "bottom", 
             annotation_legend_side = "bottom", 
             merge_legend = TRUE)
    draw(pd, x = unit(0.5, "npc"), y = unit(0.02, "npc"), 
             just = c("center", "bottom"))
  })
}
```

Cross-platform comparison

```{r fig.width=8, fig.height=4}
cell_df_filtered$cell_type <- factor(
  cell_df_filtered$cell_type,
  levels = c("B cells", "T cells", "GC B cells", "Plasma cells",
             "Macrophages", "Neutrophils", "Erythrocyte-like", 
             "Dendritic cells", "Endothelial cells", "NK cells",
             "Fibroblastic reticular cells")
)
area_heatmap_plat <- plot_area_heatmap_across_plat(cell_df_filtered, log2_scale = FALSE)

pdf("~/data/figures/area_heatmap_across_plat.pdf", width = 8, height = 4)
ComplexHeatmap::draw(area_heatmap_plat)
dev.off()
```

Cross-segmentation comparison

```{r fig.width=8, fig.height=5}
area_heatmap_seg <- plot_area_heatmap_across_seg(cell_df_filtered, log2_scale = FALSE)
grid.newpage()
grid.draw(area_heatmap_seg)

pdf("~/data/figures/area_heatmap_across_seg.pdf", width = 8, height = 5)
grid.newpage()
grid.draw(area_heatmap_seg)
dev.off()
```


#### All-in-one

```{r fig.width=15, fig.height=4}
layout <- "
AAAAABB
"

cell_area_summary_filtered <- wrap_plots(list(
  cell_area_filtered,
  median_cell_area_filtered
)) + 
  plot_layout(design = layout)
ggsave("~/data/figures/cell_area_summary_filtered.pdf", plot = cell_area_summary_filtered, width = 15, height = 4)

cell_area_summary_filtered
```

### Aspect ratio

# Correlations between metrics of interest

## Transcript counts vs cell area

# Dimensional reduction plot

Colour panel

```{r}
cols <- c("T cells" = "yellowgreen", 
          "Erythrocyte-like" = "violet", 
          "Neutrophils" = "mediumblue", 
          "B cells" = "tomato", 
          "Plasma cells" = "thistle1", 
          "GC B cells" = "yellow", 
          "Macrophages" = "skyblue2",
          "Endothelial cells" = "orange",
          "Muscle cells" = "deeppink",
          "Unknown" = "grey",
          "Dendritic cells" = "darkgreen",
          "NK cells" = "dodgerblue3",
          "Fibroblastic reticular cells" = "tan3")
```

## MERSCOPE

UMAP plots on different segmentation methods

```{r fig.width=12, fig.height=4}
merscope_dim_plots <- combine_DimPlot(
  list(merscope_vizgen, merscope_proseg, merscope_cellpose),
  cols = cols
)
ggsave("~/data/figures/merscope_dim_plots.pdf", plot = merscope_dim_plots, width = 12, height = 4)
merscope_dim_plots
```

## Xenium

Unimodal

```{r fig.width=9, fig.height=4}
xenium_uni_dim_plots <- combine_DimPlot(
  list(xenium_uni_default, xenium_uni_proseg),
  cols = cols
)
ggsave("~/data/figures/xenium_uni_dim_plots.pdf", plot = xenium_uni_dim_plots, width = 9, height = 4)
xenium_uni_dim_plots
```

Multimodal

```{r fig.width=12, fig.height=4}
xenium_multi_dim_plots <- combine_DimPlot(
  list(xenium_multi_default, xenium_multi_proseg, xenium_multi_cellpose),
  cols = cols
)
ggsave("~/data/figures/xenium_multi_dim_plots.pdf", plot = xenium_multi_dim_plots, width = 12, height = 4)
xenium_multi_dim_plots
```

# Image dimensional plot

## WT710

### Original scale with zoom-in outline

MERSCOPE

```{r fig.width=12, fig.height=4}
rect_bounds <- list(ymin=600, ymax=1400, xmin=5300, xmax=6100)

merscope_wt710_imagedimplt <- combine_ImageDimPlot(
  list(merscope_vizgen, merscope_proseg, merscope_cellpose),
  fov = "wt710",
  cols = cols,
  rect_bounds = rect_bounds
)
ggsave("~/data/figures/merscope_wt710_imagedimplt.pdf", plot = merscope_wt710_imagedimplt, width = 12, height = 4)
merscope_wt710_imagedimplt
```

Xenium unimodal

```{r fig.width=9, fig.height=4}
rect_bounds <- list(xmin = 2650, xmax = 3450, ymin = 1000, ymax = 1800)

xenium_uni_ko167_imagedimplt <- combine_ImageDimPlot(
  list(xenium_uni_default, xenium_uni_proseg),
  fov = "ko167",
  cols = cols,
  rect_bounds = rect_bounds
)
ggsave("~/data/figures/xenium_uni_ko167_imagedimplt.pdf", plot = xenium_uni_ko167_imagedimplt, width = 9, height = 4)
xenium_uni_ko167_imagedimplt
```

Xenium multimodal

```{r fig.width=9, fig.height=4}
rect_bounds <- list(xmin = 2150, xmax = 2950, ymin = 2350, ymax = 3150)

xenium_multi_wt710_imagedimplt <- combine_ImageDimPlot(
  list(xenium_multi_default, xenium_multi_proseg, xenium_multi_cellpose),
  fov = "wt710",
  cols = cols,
  rect_bounds = rect_bounds
)
ggsave("~/data/figures/xenium_multi_wt710_imagedimplt.pdf", plot = xenium_multi_wt710_imagedimplt, width = 12, height = 4)
xenium_multi_wt710_imagedimplt
```

### ROI 1: Zoom-in crop with further zoom-in outline

MERSCOPE

```{r fig.width=12, fig.height=4}
crop_bounds <- list(ymin=600, ymax=1400, xmin=5300, xmax=6100)
rect_bounds <- list(xmin = 5400, xmax = 5700, ymin = 650, ymax = 950)

merscope_wt710_imagedimplt_roi1 <- combine_ImageDimPlot(
  list(merscope_vizgen, merscope_proseg, merscope_cellpose),
  fov = "wt710",
  cols = cols,
  rect_bounds = rect_bounds,
  crop_bounds = crop_bounds,
  boundaries = "segmentation"
)
ggsave("~/data/figures/merscope_wt710_imagedimplt_roi1.pdf", plot = merscope_wt710_imagedimplt_roi1, width = 12, height = 4)
merscope_wt710_imagedimplt_roi1
```

Xenium unimodal

```{r fig.width=9, fig.height=4}
crop_bounds <- list(xmin = 2650, xmax = 3450, ymin = 1000, ymax = 1800)
rect_bounds <- list(xmin = 2750, xmax = 3050, ymin = 1050, ymax = 1350)

xenium_uni_ko167_dimplt_roi1 <- combine_ImageDimPlot(
  list(xenium_uni_default, xenium_uni_proseg),
  fov = "ko167",
  cols = cols,
  crop_bounds = crop_bounds,
  rect_bounds = rect_bounds,
  boundaries = "segmentation"
)
ggsave("~/data/figures/xenium_uni_ko167_dimplt_roi1.pdf", plot = xenium_uni_ko167_dimplt_roi1, width = 9, height = 4)
xenium_uni_ko167_dimplt_roi1
```

Xenium multimodal

```{r fig.width=12, fig.height=4}
crop_bounds <- list(xmin = 2150, xmax = 2950, ymin = 2350, ymax = 3150)
rect_bounds <- list(xmin = 2250, xmax = 2550, ymin = 2400, ymax = 2700)

xenium_multi_wt710_dimplt_roi1 <- combine_ImageDimPlot(
  list(xenium_multi_default, xenium_multi_proseg, xenium_multi_cellpose),
  fov = "wt710",
  cols = cols,
  rect_bounds = rect_bounds,
  crop_bounds = crop_bounds,
  boundaries = "segmentation"
)
ggsave("~/data/figures/xenium_multi_wt710_dimplt_roi1.pdf", plot = xenium_multi_wt710_dimplt_roi1, width = 12, height = 4)
xenium_multi_wt710_dimplt_roi1
```

### ROI 2: Further zoom-in crop for visualisation of cell polygons

MERSCOPE

```{r fig.width=12, fig.height=4}
crop_bounds <- list(xmin = 5400, xmax = 5700, ymin = 650, ymax = 950)

merscope_wt710_imagedimplt_roi2 <- combine_ImageDimPlot(
  list(merscope_vizgen, merscope_proseg, merscope_cellpose),
  fov = "wt710",
  cols = cols,
  crop_bounds = crop_bounds,
  boundaries = "segmentation"
)
ggsave("~/data/figures/merscope_wt710_imagedimplt_roi2.pdf", plot = merscope_wt710_imagedimplt_roi2, width = 12, height = 4)
merscope_wt710_imagedimplt_roi2
```

Xenium unimodal

```{r fig.width=9, fig.height=4}
crop_bounds <- list(xmin = 2750, xmax = 3050, ymin = 1050, ymax = 1350)

xenium_uni_ko167_imagedimplt_roi2 <- combine_ImageDimPlot(
  list(xenium_uni_default, xenium_uni_proseg),
  fov = "ko167",
  cols = cols,
  crop_bounds = crop_bounds,
  boundaries = "segmentation"
)
ggsave("~/data/figures/xenium_uni_ko167_imagedimplt_roi2.pdf", plot = xenium_uni_ko167_imagedimplt_roi2, width = 9, height = 4)
xenium_uni_ko167_imagedimplt_roi2
```

Xenium multimodal

```{r fig.width=12, fig.height=4}
crop_bounds <- list(xmin = 2250, xmax = 2550, ymin = 2400, ymax = 2700)

xenium_multi_wt710_imagedimplt_roi2 <- combine_ImageDimPlot(
  list(xenium_multi_default, xenium_multi_proseg, xenium_multi_cellpose),
  fov = "wt710",
  cols = cols,
  crop_bounds = crop_bounds,
  boundaries = "segmentation"
)
ggsave("~/data/figures/xenium_multi_wt710_imagedimplt_roi2.pdf", plot = xenium_multi_wt710_imagedimplt_roi2, width = 12, height = 4)
xenium_multi_wt710_imagedimplt_roi2
```

# Cell type composition

```{r}
celltype_prop <- cell_df %>%
  # filter(platform_version != "Xenium multimodal") %>% 
  dplyr::count(platform_version, segmentation_clean, sample, cell_type) %>%   # gives column `n` = count per cell_type
  group_by(platform_version, segmentation_clean, sample) %>%
  mutate(prop = n / sum(n), # sum(n) now sums across cell_type within segmentation + sample
         group = case_when(
           grepl("ctrl", sample) ~ "ctrl",
           grepl("ko", sample) ~ "ko",
           grepl("wt", sample) ~ "wt",
           TRUE ~ "other"
         )) %>% 
  ungroup()

celltype_prop$sample_id <- gsub("^[A-Za-z]+", "", celltype_prop$sample)
celltype_prop$sample_id <- factor(celltype_prop$sample_id)
celltype_prop$segmentation_clean <- factor(
  celltype_prop$segmentation_clean,
  levels = c("Default", "Proseg", "Cellpose2")
)
celltype_prop$platform_version <- factor(
  celltype_prop$platform_version,
  levels = c("MERSCOPE", "Xenium unimodal", "Xenium multimodal"))

# check if cell proportions of the same sample sum up to 1
celltype_prop %>% 
  group_by(platform_version, segmentation_clean, sample) %>% 
  summarise(sum = sum(prop), .groups = "drop")
```

## Bar chart

```{r fig.width=10, fig.height=8}
celltype_composition <- ggplot(
  celltype_prop, 
  aes(x = segmentation_clean, y = prop, fill = cell_type)
) +
  geom_col() + 
  scale_fill_manual(values = cols) + 
  facet_grid(rows = vars(platform_version), cols = vars(sample_id)) + 
  labs(x = "Segmentation",
       y = "Cell proprotion",
       fill = "Cell type") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("~/data/figures/celltype_composition.pdf", plot = celltype_composition, width = 10, height = 8, dpi = 300)
celltype_composition
```

```{r fig.width=10, fig.height=8}
celltype_composition_by_platform <- ggplot(
  celltype_prop, 
  aes(x = platform_version, y = prop, fill = cell_type)
) +
  geom_col() + 
  scale_fill_manual(values = cols) + 
  facet_grid(rows = vars(segmentation_clean), cols = vars(sample_id)) + 
  labs(x = "Segmentation",
       y = "Cell proprotion",
       fill = "Cell type") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("~/data/figures/celltype_composition_by_platform.pdf", plot = celltype_composition_by_platform, width = 10, height = 8, dpi = 300)
celltype_composition_by_platform
```

## Bar plot

```{r fig.width=6, fig.height=12}
ggplot(
  celltype_prop,
  aes(x = segmentation_clean, y = prop, fill = group)
) +
  # Bar at median
  stat_summary(
    fun = median,
    geom = "bar",
    position = position_dodge(width = 0.8),
    width = 0.8,
    alpha = 0.7
  ) +
  # Error bars (Q1–Q3 IQR)
  stat_summary(
    fun.min = function(x) quantile(x, 0.25),
    fun.max = function(x) quantile(x, 0.75),
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.3,
    linewidth = 0.5,
    color = "black"
  ) +
  # Individual data dots
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    size = 1.5,
    alpha = 0.6,
    color = "black"
  ) +
  # Median crossbar line on top
  stat_summary(
    fun = median,
    geom = "crossbar",
    position = position_dodge(width = 0.8),
    width = 0.8,
    linewidth = 0.5,
    color = "black"
  ) +
  facet_grid(cols = vars(platform_version), rows = vars(cell_type)) +
  theme_bw()
```

```{r fig.width=14, fig.height=6}
celltype_prop_bar_plot <- ggplot(
  celltype_prop %>% 
    mutate(
      platform_version = factor(
        platform_version, 
        levels = c("MERSCOPE", "Xenium multimodal", "Xenium unimodal")
      ),
      group = case_when(
        grepl("ctrl", group) ~ "CTRL",
        grepl("ko", group) ~ "KO",
        grepl("wt", group) ~ "WT"
      )
    ),
  aes(x = prop, y = segmentation_clean, fill = group)
) +
  # Bar at median
  stat_summary(
    fun = median,
    geom = "bar",
    position = position_dodge(width = 0.8),
    width = 0.8,
    alpha = 0.8
  ) +
  # Error bars (Q1–Q3 IQR)
  stat_summary(
    fun.min = function(x) quantile(x, 0.25),
    fun.max = function(x) quantile(x, 0.75),
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.3,
    linewidth = 0.5,
    color = "black"
  ) +
  # Individual data dots
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    size = 1.5,
    alpha = 0.6,
    color = "black"
  ) +
  # Median crossbar line on top
  stat_summary(
    fun = median,
    geom = "crossbar",
    position = position_dodge(width = 0.8),
    width = 0.8,
    linewidth = 0.5,
    color = "black"
  ) +
  scale_fill_manual(values = c("CTRL" = "#9B72B5", "KO" = "#D4A017", "WT" = "#3AAFA9")) +
  facet_grid(rows = vars(platform_version), cols = vars(cell_type), labeller = label_wrap_gen(width = 20)) +
  labs(x = "Cell proportion", y = "Segmentation", fill = "Group") + 
  theme_bw()
ggsave("~/data/figures/celltype_prop_bar_plot.pdf", width = 14, height = 6)
celltype_prop_bar_plot
```



```{r}
ggplot(
  celltype_prop %>% filter(platform_version == "Xenium unimodal"),
  aes(x = prop, y = cell_type, fill = group)
) +
  # Bar at median
  stat_summary(
    fun = median,
    geom = "bar",
    position = position_dodge(width = 0.8),
    width = 0.8,
    alpha = 0.7
  ) +
  # Individual data dots
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
    size = 1.5,
    alpha = 0.6,
    color = "black"
  ) +
  # Median crossbar line on top
  stat_summary(
    fun = median,
    geom = "crossbar",
    position = position_dodge(width = 0.8),
    width = 0.8,
    linewidth = 0.5,
    color = "black"
  ) +
  facet_wrap(~segmentation_clean) +
  theme_bw()
```



## Differential abundance testing with Propeller

```{r}
suppressPackageStartupMessages({
  library(speckle)
  library(scater)
  library(limma)
  library(edgeR)
  library(statmod)
  library(xtable)
})
```

```{r}
# prepare data for speckle based on segmentation method
convertPropList <- function(celltype_prop, platform_ver, seg) {
  segmentation_prop <- celltype_prop %>% 
    # only keep the current segmentation proportions and drop unwanted control group
    filter(
      platform_version == platform_ver,
      segmentation_clean == seg,
      group != "ctrl"
    ) %>% 
    select(-platform_version, -segmentation_clean)

  # make proportion data.frame with cell types as rows, samples as columns and proportions as values
  prop_wide <- segmentation_prop %>% 
    select(cell_type, sample, prop) %>% 
    pivot_wider(names_from = sample, values_from = prop) %>% 
    as.data.frame()
  rownames(prop_wide) <- prop_wide$cell_type
  prop_wide$cell_type <- NULL
  
  # make sample_info data.frame with groups and sample id as columns
  sample_info <- segmentation_prop %>% 
    select(group, sample_id) %>% 
    distinct() %>% # remove duplication
    as.data.frame()
  
  # make total_cells vector as a named vector of total cell counts by samples
  total_cells <- segmentation_prop %>% 
    group_by(sample) %>% 
    summarise(total_cells = sum(n)) %>% 
    tibble::deframe()
  
  # combine 
  prop_list <- list(
    proportions = prop_wide,
    sample_info = sample_info,
    total_cells = total_cells,
    segmentation = seg,
    platform_version = platform_ver
  )
  
  return(prop_list)
}

# test differential cell abundance across tissue types
prop_test <- function(prop_list) {
  prop <- prop_list$proportions
  sample_info <- prop_list$sample_info
  total_cells <- prop_list$total_cells
  
  # take logit transformation on cell proportions due to small sample size of 6
  logit_prop_list <- convertDataToList(
  prop, 
  data.type = "proportions", 
  transform = "logit", 
  scale.fac = total_cells
)
  
  # test WT vs KO
  designAS <- model.matrix(~0 + sample_info$group)
  colnames(designAS) <- c("KO","WT")
  
  mycontr <- makeContrasts(KO-WT, levels = designAS)
  # use t test due to two groups otherwise ANOVA
  results <- propeller.ttest(prop.list = logit_prop_list, design = designAS, 
                             contrasts = mycontr, robust = TRUE, 
                             trend = FALSE, sort = TRUE)
  
  # restuls <- results %>% 
  #   as.data.frame() %>% 
  #   tibble::rownames_to_column("cell_type") %>% 
  #   mutate(
  #     segmentation = prop_list$segmentation,
  #     platform_version = prop_list$platform_version
  #   )
  return(results)
}

plotTestResult <- function(prop_list, platform_version, segmentation, celltypes = c("B cells", "GC B cells")) {
  
  par(mfrow = c(1, length(celltypes)))
  
  for (ct in celltypes) {
    stripchart(
      as.numeric(prop_list$proportions[ct, ]) ~ prop_list$sample_info$group,
      vertical = TRUE,
      pch = c(8, 16),
      method = "jitter",
      col = c("hotpink", "darkblue"),
      cex = 2,
      ylab = "Proportions",
      group.names = c("KO", "WT"),
      main = paste(
        ct,
        segmentation,
        platform_version,
        sep = " | "
      )
    )
  }
}
```

```{r}
plat_levels <- c("MERSCOPE", "Xenium unimodal")
seg_levels <- levels(celltype_prop$segmentation_clean)

valid_combo <- celltype_prop %>% 
  filter(platform_version != "Xenium multimodal") %>% 
  distinct(platform_version, segmentation_clean)

prop_res <- valid_combo %>%
  mutate(
    prop_list = pmap(
      list(platform_version, segmentation_clean),
      function(platform_version, segmentation_clean) {
        convertPropList(
          celltype_prop,
          platform_ver = platform_version,
          seg = segmentation_clean
        )
      }
    ),
    prop_test = map(
      prop_list,
      ~ if (is.null(.x)) NULL else prop_test(.x)
    )
  )
```

```{r}
purrr::pwalk(
  list(
    prop_list          = prop_res$prop_list,
    platform_version   = prop_res$platform_version,
    segmentation_clean = prop_res$segmentation_clean
  ),
  function(prop_list, platform_version, segmentation_clean) {
    plotTestResult(
      prop_list          = prop_list,
      platform_version   = platform_version,
      segmentation = segmentation_clean
    )
  }
)

```

Cell types with significant differential abundance

```{r}
signif_prop_res <- prop_res %>%
  # unnest the prop_test list-column
  mutate(prop_df = map(prop_test, ~ as.data.frame(.x) %>% 
                       tibble::rownames_to_column(var = "Cell type"))) %>%
  select(platform_version, segmentation_clean, prop_df) %>%
  unnest(prop_df) %>%
  filter(FDR < 0.05) %>%  
  # add Segmentation column for clarity
  dplyr::rename(`Platform version` = platform_version, Segmentation = segmentation_clean)
signif_prop_res
```

```{r}
print(
  xtable(
    signif_prop_res,
    display = c("s", "s", "s", "s", rep("f", 4), "E", "E"),
    digits = c(0, 0, 0, 0, rep(4, 4), -3, -3),
    caption = "Significant differential cell proportions between WT and KO samples using Propeller",
    label = "tab:signif_differential_proportion"
  ),
  type = "latex",
  file = "~/data/signif_differential_prop.tex",
  include.rownames = FALSE,
  booktabs = TRUE
)
```

#### Heatmap

```{r}
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(grid)
})

plot_diff_abundance_heatmap <- function(prop_res, celltypes = c("B cells", "GC B cells")) {
  # prepare plot data
  plot_df <- prop_res %>%
  # unnest the prop_test list-column
  mutate(prop_df = map(prop_test, ~ as.data.frame(.x) %>% 
                       tibble::rownames_to_column(var = "Cell type"))) %>%
  select(platform_version, segmentation_clean, prop_df) %>%
  unnest(prop_df) %>%
  filter(`Cell type` %in% celltypes) %>%  
  # add Segmentation column for clarity
  dplyr::rename(`Platform version` = platform_version, Segmentation = segmentation_clean) %>% 
  ungroup()
  
  segmentations <- c("Default", "Proseg", "Cellpose2")
  platform_versions <- c("MERSCOPE", "Xenium unimodal")
  segmentation_colours <- c(
    "Default" = "#F8766D",
    "Proseg" = "#00BA38",
    "Cellpose2" = "#619CFF"
  )
  
  # build one matrix per platform (rows = platforms, cols = segmentation x cell_pair)
  col_order <- expand.grid(segmentation = segmentations, cell_type = celltypes, stringsAsFactors = FALSE)
  col_order <- col_order[order(match(col_order$cell_type, celltypes),
                               match(col_order$segmentation, segmentations)), ]
  col_labels <- paste(col_order$cell_type, col_order$segmentation, sep = "|")
  mat <- matrix(NA, nrow = length(platform_versions), ncol = nrow(col_order), 
                dimnames = list(platform_versions, col_labels))
  sig_mat <- matrix(NA, nrow = length(platform_versions), ncol = nrow(col_order), 
                dimnames = list(platform_versions, col_labels))
  
  for (i in seq_len(nrow(plot_df))) {
    r <- plot_df$`Platform version`[i]
    c <- paste(plot_df$`Cell type`[i], plot_df$Segmentation[i], sep = "|")
    mat[r, c] <- plot_df$PropRatio[i]
    sig_mat[r, c] <- plot_df$FDR[i] < 0.05
  }
  
  # set annotation for segmentation colour bar
  seg_bar <- HeatmapAnnotation(
    Segmentation = col_order$segmentation,
    col = list(Segmentation = segmentation_colours),
    show_annotation_name = TRUE,
    annotation_name_side = "right",
    annotation_name_gp = gpar(fontsize = 10),
    simple_anno_size = unit(5, "mm"),
    annotation_legend_param = list(
      Segmentation = list(
        title = "Segmentation",
        at = segmentations,
        labels = segmentations,
        nrow = 1,
        legend_gp = gpar(fill = segmentation_colours)
      )
    )
  )

  # set colour scale
  col_fun <- colorRamp2(
    c(0, 0.5, 1),
    c("#08306B", "#6BAED6", "white")
  )
  
  # column split by cell pair
  col_split <- factor(col_order$cell_type, levels = celltypes)
  
  # draw heatmap
  ht_main <- Heatmap(
    mat,
    name = "Mean proportion ratio",
    col = col_fun,
    
    # column settings
    column_split = col_split,
    column_gap = unit(1.5, "mm"),
    column_title_gp   = gpar(fontsize = 10),
    column_title_side = "top",            # segmentation labels at top
    show_column_names = FALSE,            # hide individual segmentation names
    column_title_rot = 45,
    cluster_columns   = FALSE,
    cluster_column_slices = FALSE,
    
    # row settings
    cluster_rows      = FALSE,
    row_names_side    = "right",
    row_names_gp      = gpar(fontsize = 10),
    
    # cell size
    width = unit(ncol(mat) * 5, "mm"),
    height = unit(nrow(mat) * 5, "mm"),
    
    # top annotation (segmentation bar below cell pair title)
    top_annotation = seg_bar,
    
    # significance stars
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (!is.na(sig_mat[i, j]) & sig_mat[i, j]) {
        grid.text("*", x, y, gp = gpar(fontsize = 10, col = "black"))
      }
    },
    
    # borders
    border            = TRUE,
    rect_gp           = gpar(col = "white", lwd = 0.5),
    
    # legend
    heatmap_legend_param = list(
      title         = "Mean proportion ratio",
      at            = seq(0, 1, length.out = 3),
      direction     = "horizontal",
      title_position = "lefttop",
      legend_width  = unit(5, "cm")
    )
  )
  
  draw(ht_main,
     heatmap_legend_side    = "right",
     annotation_legend_side = "right",
     merge_legend           = TRUE)
}
```

```{r fig.width=6, fig.height=2}
cell_abundance_ht <- plot_diff_abundance_heatmap(prop_res)

# save
pdf("~/data/figures/cell_diff_abundance_heatmap.pdf", width = 6, height = 2)
cell_abundance_ht
dev.off()
```

Heat map with p-values

```{r}
plot_diff_abundance_heatmap2 <- function(prop_res, celltypes = c("B cells", "GC B cells")) {
  # prepare plot data
  plot_df <- prop_res %>%
  # unnest the prop_test list-column
  mutate(prop_df = map(prop_test, ~ as.data.frame(.x) %>% 
                       tibble::rownames_to_column(var = "Cell type"))) %>%
  select(platform_version, segmentation_clean, prop_df) %>%
  unnest(prop_df) %>%
  filter(`Cell type` %in% celltypes) %>%  
  # add Segmentation column for clarity
  dplyr::rename(`Platform version` = platform_version, Segmentation = segmentation_clean) %>% 
  ungroup() %>% 
  mutate(
    # formatted labels
    PropRatio_label = sprintf("%.2f", PropRatio),
    adj_p_value_label = case_when(
      FDR < 0.001 ~ paste0("p=", formatC(FDR, format = "e", digits = 1), " *"),
      FDR < 0.05 ~ paste0("p=", sprintf("%.3f", FDR), " *"),
      TRUE ~ paste0("p=", sprintf("%.3f", FDR))
    ),
    tile_label = paste(PropRatio_label, adj_p_value_label, sep = "\n"),
    text_color = if_else(
      PropRatio < 0.5, 
      "white",
      "black"
    ),
    text_face = if_else(
      FDR < 0.05,
      "bold",
      "plain"
    )
  )
  
  # heat map
  ggplot(plot_df, aes(x = Segmentation, y = `Cell type`, fill = PropRatio)) + 
    geom_tile(aes(color = "white"), linewidth = 1) + 
    geom_text(aes(label = tile_label, color = text_color, fontface = text_face), 
              size = 3.5, lineheight = 1) + 
    scale_colour_manual(
      values = c("white" = "white", "black" = "black"),
      guide = "none"   # hides legend
    ) + 
    scale_fill_gradientn(
      colours = c("darkblue", "blue", "royalblue", "lightskyblue", "white"),
      limits = c(0, 1),
      oob = scales::squish,
      na.value = "grey80",
      name = "Mean proportion ratio",
      guide = guide_colorbar(
        title.position = "top",    # "top", "bottom", "left", "right"
        # label.position = "left",   # "left", "right", "top", "bottom"
        title.hjust = 0.5          # centre the title
      )
    ) + 
    facet_grid(. ~`Platform version`, scales = "free_x", space = "free_x") + 
    labs(
      x = "Segmentation",
      y = "Cell type"
    ) + 
    theme_bw() + 
    theme(
      # axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom",
      panel.grid = element_blank()
    )
}
```

```{r fig.width=6, fig.height=3}
cell_abundance_ht2 <- plot_diff_abundance_heatmap2(prop_res)
ggsave("~/data/figures/cell_diff_abundance_heatmap_pvalue.pdf", plot = cell_abundance_ht2, width = 6, height = 3)
cell_abundance_ht2
```


# Pseudo-bulk DE analysis

Read the result table

```{r}
res_all <- read.csv("~/SpatialBench_segmentation_comparison/Data/pseudo-bulk_DE_results.csv") %>% 
  filter(group == "KOvsWT")
res_all$segmentation <- factor(
  res_all$segmentation,
  levels = c("Default", "Proseg", "Cellpose2")
)
```

## Volcano plots on selected cell types

Function

```{r}
# volcano plots showing DE calls across segmentation methods and platforms for a given cell type
celltype_volcano_plot <- function(de_res) {
  cell_type <- unique(de_res$cell_type)
  ggplot(
    de_res, 
    aes(x = logFC, y = -log10(adj.P.Val), color = Status)
) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = c(
    "Significant (Downregulated)" = "orchid3",
    "Significant (Upregulated)" = "skyblue3",
    "False Significant" = "red",
    "Not Significant" = "black"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "darkred") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  scale_x_continuous(limits = c(-4.5, 4.5), breaks = seq(-4, 4, 2)) +
  scale_y_continuous(limits = c(0, 5)) + 
  labs(
    title = cell_type,
    x = "Log2 Fold Change",
    y = "-Log10(adj P-value)",
    color = "DE Status"
  ) +
  facet_grid(rows = vars(platform), cols = vars(segmentation)) +
  theme_bw() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 12),
    legend.position = "bottom"
  ) +
  geom_label_repel(
    # data = subset(de_res, Status %in% c("Significant (Downregulated)", "Significant (Upregulated)")),
    data = subset(de_res, gene %in% c("Tbx21", "Cxcr3") & Status == "Significant (Downregulated)"),
    aes(label = gene),
    box.padding = 0.35,
    point.padding = 0.5,
    point.size = 0.3,
    min.segment.length = 0, # uncomment to force drawing the segment from label to the point
    segment.color = "black",
    show.legend = FALSE,
    size = 3.5
  )
}
```

Select B cells, GC B cells, and T cells on the kovswt comparison

```{r fig.width=9, fig.height=6}
res_all_b_cell <- res_all %>% 
  filter(cell_type == "B cells")

volcano_plot_b_cell <- celltype_volcano_plot(res_all_b_cell)
volcano_plot_b_cell
```

```{r fig.width=9, fig.height=6}
res_all_gc_b_cell <- res_all %>% 
  filter(cell_type == "GC B cells")

volcano_plot_gc_b_cell <- celltype_volcano_plot(res_all_gc_b_cell)
volcano_plot_gc_b_cell
```

```{r fig.width=9, fig.height=6}
res_all_t_cell <- res_all %>%
  filter(cell_type == "T cells")

volcano_plot_t_cell <- celltype_volcano_plot(res_all_t_cell)
volcano_plot_t_cell
```

```{r fig.width=7, fig.height=9}
volcano_plot <- (volcano_plot_b_cell + NoLegend()) / volcano_plot_gc_b_cell
ggsave("~/data/figures/volcano_plot.pdf", plot = volcano_plot, width = 7, height = 9)
volcano_plot
```

## Heat map on selected gene

Function

```{r}
# heat map showing differential gene expression across cell types, segmentation methods, and platforms for a given gene
celltype_de_heatmap <- function(de_res, gene_use) {
  # all combinations that should appear in the heatmap
  full_grid <- de_res %>% 
    distinct(cell_type, segmentation, platform)
  
  # subset to selected gene
  de_res_sub <- de_res %>% 
    filter(gene == gene_use)
  
  # join to full grid (introduces NA where gene is missing)
  plot_df <- full_grid %>% 
    left_join(de_res_sub, 
              by = c("cell_type", "segmentation", "platform")) %>% 
    mutate(
      # formatted labels
      logFC_label = if_else(
        is.na(logFC), 
        NA_character_, 
        sprintf("%.2f", logFC)
      ),
      adj_p_val_label = case_when(
        is.na(adj.P.Val) ~ NA_character_,
        adj.P.Val < 0.001 ~ paste0("p=", formatC(adj.P.Val, format = "e", digits = 1), " 
                                   *"),
        adj.P.Val < 0.05 ~ paste0("p=", sprintf("%.3f", adj.P.Val), " *"),
        TRUE ~ paste0("p=", sprintf("%.3f", adj.P.Val))
      ),
      tile_label = if_else(
        is.na(logFC),
        "",
        paste(logFC_label, adj_p_val_label, sep = "\n")
      ),
      text_color = if_else(
        !is.na(logFC) & logFC < -2, 
        "white",
        "black"
      ),
      # text_color = case_when(
      #   !is.na(adj.P.Val) & adj.P.Val < 0.05 ~ "darkorange",
      #   TRUE ~ text_color
      # ), 
      text_face = if_else(
        !is.na(adj.P.Val) & adj.P.Val < 0.05,
        "bold",
        "plain"
      )
    )
  
  # heat map
  ggplot(plot_df, aes(x = segmentation, y = cell_type, fill = logFC)) + 
    geom_tile(aes(color = "white"), linewidth = 1) + 
    geom_text(aes(label = tile_label, color = text_color, fontface = text_face), 
              size = 3.5, lineheight = 1) + 
    scale_colour_manual(
      values = c("white" = "white", "black" = "black", "darkorange" = "darkorange"),
      guide = "none"   # hides legend
    ) + 
    scale_fill_gradient2(
      low = "darkblue", mid = "white", high = "darkred", 
      midpoint = 0, 
      limits = c(-4.5, 1), oob = scales::squish,
      na.value = "grey80", name = "Log2FC"
    ) + 
    facet_grid(. ~platform, scales = "free_x", space = "free_x") + 
    labs(
      title = paste("Differential expression of", gene_use),
      x = "Segmentation",
      y = "Cell type"
    ) + 
    theme_bw() + 
    theme(
      # axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    )
}
```

```{r fig.width=9, fig.height=6}
tbx21_de_heatmap <- celltype_de_heatmap(res_all, gene_use = "Tbx21")
ggsave("~/data/figures/tbx21_de_heatmap.pdf", plot = tbx21_de_heatmap, width = 9, height = 6)
tbx21_de_heatmap
```

# Purity analysis

# MECR

Read MECR information

```{r}
MECR_sp <- readRDS("~/SpatialBench_segmentation_comparison/Data/MECR_info_customised.rds")
MECR_sc <- readRDS("~/SpatialBench_segmentation_comparison/Data/MECR_sc.rds")
```

Merge across segmentation and platforms

```{r}
MECR_df <- imap_dfr(MECR_sp, ~{
  parts <- strsplit(.y, "_")[[1]]
  .x[["MECR_df"]] %>% 
    mutate(
      # standardise the order of cell pairs
      cell_pair = paste0(
        "(", 
        pmin(cell_type1, cell_type2), 
        ", ", 
        pmax(cell_type1, cell_type2), 
        ")"
      ),
      platform_version = parts[1],
      segmentation = parts[2]
    )
})
MECR_df$segmentation <- factor(
  MECR_df$segmentation,
  levels = c("Default", "Proseg", "Cellpose2")
)
# count the number of gene pairs under each cell pair
MECR_df <- MECR_df %>% 
  group_by(platform_version, segmentation, cell_pair) %>% 
  mutate(n_pairs = n()) %>% 
  ungroup()

MECR_df_filtered <- MECR_df %>% 
  filter(platform_version != "Xenium unimodal")
head(MECR_df_filtered)
```

```{r}
MECR_sc_df <- MECR_sc$MECR_df %>% 
  mutate(
    # standardise the order of cell pairs
    cell_pair = paste0(
      "(", 
      pmin(cell_type1, cell_type2), 
      ", ", 
      pmax(cell_type1, cell_type2), 
      ")"
    )
  )
head(MECR_sc_df)
```

### Segmentation level

Filter out the B and GC B gene pairs as GC B cells can express some B cell markers.

```{r}
MECR_df_filtered <- MECR_df_filtered %>% 
  filter(
    # n_pairs > 10,
    cell_pair != "(B cells, GC B cells)"
    )

MECR_sc_df_filtered <- MECR_sc_df %>% 
  group_by(cell_pair) %>% 
  mutate(n_pairs = n()) %>% 
  filter(
    # n_pairs > 10,
    cell_pair != "(B cells, GC B cells)") %>% 
  ungroup()
```


```{r fig.width=5, fig.height=4}
MECR_seg_vlnplt <- ggplot(
    MECR_df_filtered, 
    aes(x = segmentation, y = MECR, fill = segmentation)
  ) + 
  geom_violin(trim = FALSE, scale = "width") + 
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white", color = "black") +
  scale_fill_manual(values = colour_panel) +
  facet_wrap(~platform_version) + 
  labs(x = "Segmentation",
       y = "MECR") + 
  scale_y_continuous(limits = c(0, 1), n.breaks = 8) + 
  theme_bw() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
ggsave("~/data/figures/MECR_seg_vlnplt.pdf", plot = MECR_seg_vlnplt, width = 5, height = 4)
MECR_seg_vlnplt
```

Compare to MECR from single cell reference 

```{r fig.width=3, fig.height=4}
MECR_sc_vlnplt <- ggplot(
    MECR_sc_df_filtered, 
    aes(x = "", y = MECR)
  ) + 
  geom_violin(trim = FALSE, scale = "width", fill = "grey") + 
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white", color = "black") +
  scale_fill_manual(values = colour_panel) +
  labs(x = "",
       y = "MECR") + 
  scale_y_continuous(limits = c(0, 1), n.breaks = 8) + 
  theme_bw() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
MECR_sc_vlnplt
```

Combine 

```{r}
MECR_vlnplt <- MECR_sc_vlnplt + MECR_seg_vlnplt + plot_layout(widths = c(1, 6))
ggsave("~/data/figures/MECR_vlnplt_ref.pdf", plot = MECR_vlnplt, width = 6, height = 4)
MECR_vlnplt
```


### Cell pair-specific

```{r}
MECR_cellpair_top <- MECR_df_filtered %>% 
  group_by(platform_version, segmentation, cell_pair) %>% 
  summarise(MECR_median = median(MECR), n_pairs = n(), .groups = "drop") %>% 
  group_by(platform_version, segmentation) %>%
  arrange(desc(MECR_median)) %>% 
  slice_head(n = 10)

# find the common cell pairs with most contamination across platforms and segmentation
# n = 8 means this cell pair is found in every platform and segmentation
MECR_cellpair_top %>% group_by(cell_pair) %>% summarise(n = n(), n_pairs = first(n_pairs), MECR_median = median(MECR_median)) %>% arrange(desc(n))
```

```{r}
MECR_cellpair_bottom <- MECR_df %>% 
  group_by(platform_version, segmentation, cell_pair) %>% 
  summarise(MECR_median = median(MECR), n_pairs = n(), .groups = "drop") %>% 
  group_by(platform_version, segmentation) %>%
  arrange(MECR_median) %>% 
  slice_head(n = 10)

# find the common cell pairs with least contamination across platforms and segmentation
MECR_cellpair_bottom %>% group_by(cell_pair) %>% summarise(n = n(), n_pairs = first(n_pairs)) %>% arrange(desc(n))
```

#### LMM (TODO) 

First, use linear mixed-effects model (LMM). Treat cell pair as a random effect and combination of platform and segmentation as fixed effects.

```{r}
suppressPackageStartupMessages({
  library(lme4)
  library(lmerTest)
})

fit <- lmer(
  MECR ~ platform_version * segmentation + (1 | cell_pair),
  data = MECR_df
)

anova(fit)
summary(fit)
```

**Interpretation:**

-   TypeIII ANOVA table:
    -   All three effects are highly significant:
        1.  Platform matters - MECR differ strongly between MERSCOPE, Xenium unimodal, and Xenium multimodal.
        2.  Segmentation matters even more - Larger F value indicates that segmentation choices have bigger impact on MECR than platform.
        3.  There is a platform x segmentation interaction - segmentation effect is not the same on all platforms, cannot interpret platform effects without considering segmentation.
-   Model structure (sanity check)
    -   5344 MECR measurements
    -   55 distinct cell pairs
    -   Each cell pair has its own baseline MECR (random intercept)
-   Random effects (how much variation comes from cell pairs)
    -   About half of the unexplained variance is due to differences between cell pairs ($\sigma^2_\text{cell_pair}/(\sigma^2_\text{cell_pair}+\sigma^2_\text{residual})$)
    -   The other half is the within-cell-pair variability across platform/segmentation.
    -   Hence cell pairs are genuinely different from each other and should model them as random effects.
-   Fixed effects
    -   Reference level - MERSCOPE + Default
        -   Intercept = mean MECR for MERSCOPE + Default segmentation
    -   Main effects: platform
        -   At default segmentation:
            -   Xenium multimodal MECR is \~0.2 higher than MERSCOPE.
            -   Xenium unimodal MECR is \~0.023 higher than MERSCOPE.
    -   Main effects: segmentation
        -   On MERSCOPE:
            -   Proseg increases MECR by \~0.21.
            -   Cellpose2 slightly decreases MECR (\~-0.024).
    -   Interaction terms (how segmentation effects change depending on platform)
        -   Xenium multimodal x Proseg (+0.064)
            -   Proseg boosts MECR even more on Xenium multimodal than on MERSCOPE.
            -   Total Proseg effect on Xenium multimodal is: 0.221 (Proseg) + 0.064 = \~0.275
        -   Xenium unimodal x Proseg (no interaction)
            -   Proseg behaves basically the same on Xenium unimodal as on MERSCOPE.
        -   Xenium multimodal x Cellpose2 (-0.042)
            -   Cellpose2 performs worse on Xenium multimodal than expected from its MERSCOPE effect.
            -   Platform-specific penalty on Cellpose2.
-   Rank deficiency warning due to missing of one paltform-segmentation combination (Xenium unimodal + Cellpose2)

**Conclusion:**

> MECR values vary strongly by both platform and segmentation, with segmentation having the largest overall effect, and with clear platform-specific behavior—especially for Proseg and Cellpose2 on Xenium multimodal—while substantial baseline differences exist between cell pairs.'

```{r}
library(emmeans)
emm <- emmeans(fit, ~ platform_version * segmentation)
plot(emm)
```

Which cell pairs differ most overall?

Based on per-cell-pair baseline shifts and deviation from the global mean.

```{r}
re <- ranef(fit)$cell_pair
re$abs_effect <- abs(re[,1])
re <- re[order(-re$abs_effect), ]
re
```

Which cell pairs show different patterns across platform/segmentation?

```{r}
fit_rs <- lmer(
  MECR ~ platform_version * segmentation +
    (platform_version * segmentation | cell_pair),
  data = MECR_df
)
ranef(fit_rs)$cell_pair
```


Find the highly variable cell pairs across segmentation and platforms.

```{r}
cellpair_var <- MECR_df %>% 
  group_by(cell_pair) %>% 
  summarise(
    MECR_sd = sd(MECR, na.rm = TRUE),
    MECR_range = max(MECR) - min(MECR)
  ) %>% 
  arrange(desc(MECR_sd))
cellpair_var
```

Select representative cell pairs for heat map

```{r}
top_cellpair <- MECR_cellpair_top %>% 
  group_by(cell_pair) %>% 
  summarise(n = n()) %>%
  group_by(n) %>% 
  slice(1) %>% 
  pull(cell_pair)

bottom_cellpair <- MECR_cellpair_bottom %>% 
  group_by(cell_pair) %>% 
  summarise(n = n()) %>%
  group_by(n) %>% 
  slice(1) %>% 
  pull(cell_pair)

variable_pairs <- head(cellpair_var$cell_pair)
```

#### Heatmap

Functions

```{r}
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(grid)
})

plot_MECR_cellpair_heatmap <- function(MECR_sp, MECR_sc, cellpair_use = NULL, show_n_pairs = TRUE, show_anno_legend = FALSE) {
  if (is.null(cellpair_use)) {
    cellpair_use <-  unique(MECR_sp$cell_pair)
  }
  # prepare plot data
  plot_df <- MECR_sp %>% 
    filter(cell_pair %in% cellpair_use) %>% 
    group_by(platform_version, segmentation, cell_pair) %>% 
    summarise(MECR_median = median(MECR), n_pairs = n(), .groups = "drop") %>% 
    ungroup()
  
  cell_pairs <- cellpair_use
  segmentations <- c("Default", "Proseg", "Cellpose2")
  platform_versions <- c("MERSCOPE", "Xenium multimodal")
  segmentation_colours <- c(
    "Default" = "#F8766D",
    "Proseg" = "#00BA38",
    "Cellpose2" = "#619CFF"
  )
  
  # build one matrix per platform (rows = platforms, cols = segmentation x cell_pair)
  col_order <- expand.grid(segmentation = segmentations, cell_pair = cell_pairs, stringsAsFactors = FALSE)
  col_order <- col_order[order(match(col_order$cell_pair, cell_pairs),
                               match(col_order$segmentation, segmentations)), ]
  col_labels <- paste(col_order$cell_pair, col_order$segmentation, sep = "|")
  mat <- matrix(NA, nrow = length(platform_versions), ncol = nrow(col_order), 
                dimnames = list(platform_versions, col_labels))
  
  for (i in seq_len(nrow(plot_df))) {
    r <- plot_df$platform_version[i]
    c <- paste(plot_df$cell_pair[i], plot_df$segmentation[i], sep = "|")
    mat[r, c] <- plot_df$MECR_median[i]
  }
  
  # set annotation for segmentation colour bar
  seg_bar <- HeatmapAnnotation(
    Segmentation = col_order$segmentation,
    col = list(Segmentation = segmentation_colours),
    show_annotation_name = TRUE,
    annotation_name_side = "right",
    annotation_name_gp = gpar(fontsize = 10),
    simple_anno_size = unit(5, "mm"),
    annotation_legend_param = list(
      Segmentation = list(
        title = "Segmentation",
        at = segmentations,
        labels = segmentations,
        nrow = 1,
        legend_gp = gpar(fill = segmentation_colours)
      )
    ),
    show_legend = show_anno_legend
  )

  # set colour scale
  col_fun <- colorRamp2(
    seq(0, 1, length.out = 5),
    c("white", "#FFE5CC", "#FFB347", "#FF8C00", "#FF6600")
  )
  
  # column split by cell pair
  col_split <- factor(col_order$cell_pair, levels = cell_pairs)
  
  # draw heatmap
  ht_main <- Heatmap(
    mat,
    name = "median MECR",
    col = col_fun,
    
    # column settings
    column_split = col_split,
    column_gap = unit(1.5, "mm"),
    column_title_gp   = gpar(fontsize = 9),
    column_title_side = "top",            # segmentation labels at top
    show_column_names = FALSE,            # hide individual segmentation names
    column_title_rot = 45,
    cluster_columns   = FALSE,
    cluster_column_slices = FALSE,
    
    # row settings
    cluster_rows      = FALSE,
    row_names_side    = "right",
    row_names_gp      = gpar(fontsize = 9),
    
    # cell size
    width = unit(ncol(mat) * 5, "mm"),
    height = unit(nrow(mat) * 5, "mm"),
    
    # top annotation (segmentation bar below cell pair title)
    top_annotation = seg_bar,
    
    # borders
    border            = TRUE,
    rect_gp           = gpar(col = "white", lwd = 0.5),
    
    # legend
    heatmap_legend_param = list(
      title         = "Median MECR",
      at            = seq(0, 1, length.out = 5),
      direction     = "horizontal",
      title_position = "lefttop",
      legend_width  = unit(5, "cm")
    )
  )
  
  # add single-cell reference MECR
  MECR_sc_median <- MECR_sc %>% 
    group_by(cell_pair) %>% 
    summarise(MECR_median = median(MECR), .groups = "drop") %>% 
    arrange(match(cell_pair, cell_pairs))
  
  mat_sc <- matrix(NA, nrow = 1, ncol = nrow(col_order),
                   dimnames = list("scRNA-seq", col_labels))
  
  for (i in seq_len(nrow(col_order))) {
    cp <- col_order$cell_pair[i]
    mat_sc[1, i] <- MECR_sc_median$MECR_median[MECR_sc_median$cell_pair == cp]
  }
  
  ht_sc <- Heatmap(
      mat_sc,
      name = "scRNA-seq",
      col  = col_fun,
  
      column_split          = col_split,        # same split as ht_main
      column_gap            = unit(1.5, "mm"),
      show_column_names     = FALSE,
      column_title          = NULL,             # no repeated labels
      cluster_columns       = FALSE,
      cluster_column_slices = FALSE,
  
      cluster_rows   = FALSE,
      row_names_side = "right",
      row_names_gp   = gpar(fontsize = 9),
  
      width  = unit(ncol(mat_sc) * 5, "mm"),
      height = unit(5, "mm"),
  
      border  = TRUE,
      rect_gp = gpar(lwd = 0),
      show_heatmap_legend = FALSE
    )
  
  if (show_n_pairs) {
    # n_pairs: one value per cell_pair (collapsed across segmentations)
    n_pairs <- plot_df %>%
      group_by(cell_pair) %>%
      summarise(n_pairs = mean(n_pairs, na.rm = TRUE), .groups = "drop") %>%
      arrange(match(cell_pair, cell_pairs))
  
    mat_n <- matrix(NA, nrow = 1, ncol = nrow(col_order),
                    dimnames = list("N pairs", col_labels))
    
    for (i in seq_len(nrow(col_order))) {
      cp <- col_order$cell_pair[i]
      mat_n[1, i] <- n_pairs$n_pairs[n_pairs$cell_pair == cp]
    }
  
    # n_pairs colour scale
    n_max <- max(mat_n, na.rm = TRUE)
    col_fun_n <- colorRamp2(
      c(0, n_max / 2, n_max),
      c("white", "#6BAED6", "#08306B")
    )
    
    # n_pairs heatmap — same column structure as ht_main
    ht_n <- Heatmap(
      mat_n,
      name = "N pairs",
      col  = col_fun_n,
  
      column_split          = col_split,        # same split as ht_main
      column_gap            = unit(1.5, "mm"),
      show_column_names     = FALSE,
      column_title          = NULL,             # no repeated labels
      cluster_columns       = FALSE,
      cluster_column_slices = FALSE,
  
      cluster_rows   = FALSE,
      row_names_side = "right",
      row_names_gp   = gpar(fontsize = 9),
  
      width  = unit(ncol(mat_n) * 5, "mm"),
      height = unit(5, "mm"),
  
      # show n value once per group (only label the middle column)
      cell_fun = function(j, i, x, y, width, height, fill) {
        if (j %% length(segmentations) == 2) {   # middle segmentation column
          value <- round(mat_n[i, j])
          text_colour <- ifelse(value > 20, "white", "black")
          grid.text(
            label = value,
            x, y,
            gp = gpar(fontsize = 8, col = text_colour)
          )
        }
      },
  
      border  = TRUE,
      rect_gp = gpar(lwd = 0),
  
      heatmap_legend_param = list(
        title          = "N pairs",
        direction      = "horizontal",
        title_position = "lefttop",
        legend_width   = unit(4, "cm")
      )
    )
    
    # combine
    ht_list <- ht_main %v% ht_sc %v% ht_n
  } else {
    ht_list <- ht_main %v% ht_sc
  }
  
  draw(ht_list,
     heatmap_legend_side    = "bottom",
     annotation_legend_side = "bottom",
     merge_legend           = TRUE)
}
```

```{r fig.width=8, fig.height=4}
top_ht <- plot_MECR_cellpair_heatmap(MECR_df_filtered, MECR_sc$MECR_df, cellpair_use = c(top_cellpair[-1], "(Erythrocyte-like, T cells)"))

# save
pdf("~/data/figures/MECR_top_cellpair_heatmap_n_pairs.pdf", width = 8, height = 4)
top_ht
dev.off()
```

```{r fig.width=8, fig.height=4}
bottom_ht <- plot_MECR_cellpair_heatmap(MECR_df, cellpair_use = bottom_cellpair)

# save
pdf("~/data/figures/MECR_bottom_cellpair_heatmap.pdf", width = 8, height = 4)
bottom_ht
dev.off()
```

```{r fig.width=8, fig.height=4}
var_ht <- plot_MECR_cellpair_heatmap(MECR_df, cellpair_use = variable_pairs)

# save
pdf("~/data/figures/MECR_variable_cellpair_heatmap_n_pairs.pdf", width = 8, height = 4)
var_ht
dev.off()
```


```{r fig.width=8, fig.height=4}
var_ht <- plot_MECR_cellpair_heatmap(MECR_df, cellpair_use = variable_pairs2)

# # save
# pdf("~/data/figures/MECR_variable_cellpair_heatmap.pdf", width = 8, height = 4)
# var_ht
# dev.off()
```

Combine segmentation and platforms

```{r fig.width=7, fig.height=4}
library(tidyr)
library(dplyr)
library(ComplexHeatmap)
library(circlize)

MECR_sc_df_filtered <- MECR_sc_df_filtered %>% 
  group_by(cell_pair) %>% 
  mutate(n_pairs = n(), method_combo = "scRNA-seq") %>% 
  ungroup()
MECR_mat <- MECR_df_filtered %>%
  unite(method_combo, platform_version, segmentation, sep = " ") %>%
  bind_rows(MECR_sc_df_filtered) %>% 
  # filter(n_pairs > 10) %>% 
  group_by(method_combo, cell_pair) %>% 
  summarise(MECR_median = median(MECR, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(
    names_from = method_combo,
    values_from = MECR_median
  ) %>%
  column_to_rownames("cell_pair") %>% 
  t()

row_order <- c("scRNA-seq", "MERSCOPE Default", "MERSCOPE Proseg", "MERSCOPE Cellpose2",
               "Xenium multimodal Default", "Xenium multimodal Proseg", "Xenium multimodal Cellpose2")

# cluster manually
mat4clust <- MECR_mat[row_order, ]
d <- dist(t(mat4clust), method = "euclidean")
hc <- hclust(d, method = "ward.D2")

col_order <- colnames(mat4clust)[hc$order]

n_pairs_mat <- MECR_df_filtered %>% 
  group_by(cell_pair) %>% 
  summarise(n_pairs = first(n_pairs)) %>% 
  # Build a 1-row matrix matching column order of mecr_matrix
  { setNames(.$n_pairs, .$cell_pair) } %>%
  .[col_order] %>%
  matrix(nrow = 1, dimnames = list("N pairs", names(.))) 

n_max <- max(n_pairs_mat, na.rm = TRUE)
col_fun_n <- colorRamp2(
  c(0, n_max / 2, n_max),
  c("white", "#6BAED6", "#08306B")
)

# --- N pairs annotation heatmap (bottom) ---
npairs_heatmap <- Heatmap(
  n_pairs_mat,
  name = "N pairs",
  col  = col_fun_n,
  cell_fun = function(j, i, x, y, width, height, fill) {
    value <- round(n_pairs_mat[i, j])
    text_colour <- ifelse(value > 20, "white", "black")
    grid.text(
      label = value,
      x, y,
      gp = gpar(fontsize = 8, col = text_colour)
    )
  },
  show_row_names    = TRUE,
  show_column_names = FALSE,   # shared with main heatmap above
  row_names_side    = "right",
  column_order      = col_order,  # lock column order
  cluster_rows      = FALSE,
  cluster_columns   = FALSE,
  width  = unit(ncol(n_pairs_mat) * 5, "mm"),
  height = unit(5, "mm"),
  heatmap_legend_param = list(
    title          = "N pairs",
    at             = c(0, 15, 30, 45, 60),
    direction      = "horizontal",
    title_position = "lefttop"
  )
)

# --- Main MECR heatmap ---
mecr_heatmap <- Heatmap(
  MECR_mat[, col_order],
  name = "Median MECR",
  col  = colorRamp2(
    seq(0, 1, length.out = 5),
    c("white", "#FFE5CC", "#FFB347", "#FF8C00", "#FF6600")
  ),
  cluster_rows      = TRUE,
  cluster_columns   = FALSE,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_side    = "right",
  column_names_side = "top",
  column_names_rot  = 45,
  column_names_gp   = gpar(fontsize = 8),
  row_names_gp      = gpar(fontsize = 10),
  rect_gp           = gpar(col = "white", lwd = 1.5),
  width = unit(ncol(MECR_mat) * 5, "mm"),
  height = unit(nrow(MECR_mat) * 5, "mm"),
  heatmap_legend_param = list(
    title          = "Median MECR",
    at             = c(0, 0.25, 0.5, 0.75, 1),
    direction      = "horizontal",
    title_position = "lefttop"
  )
)

# --- Draw stacked ---
pdf("~/data/figures/MECR_heatmap_filtered_n_pairs.pdf", width = 7, height = 4)
draw(
  mecr_heatmap %v% npairs_heatmap,   # %v% stacks vertically
  heatmap_legend_side   = "bottom",
  annotation_legend_side = "bottom",
  merge_legends = TRUE
)
dev.off()

draw(
  mecr_heatmap %v% npairs_heatmap,   # %v% stacks vertically
  heatmap_legend_side   = "bottom",
  annotation_legend_side = "bottom",
  merge_legends = TRUE
)
```


```{r fig.width=15, fig.height=4}
draw(
  mecr_heatmap %v% npairs_heatmap,   # %v% stacks vertically
  heatmap_legend_side   = "bottom",
  annotation_legend_side = "bottom",
  merge_legends = TRUE
)
```



