---
title: "pseudo_bulk.rmd"
output: 
  html_document:
    toc: true
    toc_float: 
      collapsed: false
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/data")
```

# Load packages

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(tidyverse)
  library(limma)
  library(edgeR)
  library(ggplot2)
  library(ggrepel)
  library(dplyr)
  library(ggpubr)
  library(scales)
  library(ComplexUpset)
  library(patchwork)
})
```

# Load datasets

```{r}
# default segmentation
merged_default <- readRDS("default_segmentation/SingleR_annotated/merged_default.rds")

# baysor segmentation
merged_baysor <- readRDS("baysor_segmentation/dimension_reduction/SingleR_annotation/merged_baysor.rds")

# cellpose2 segmentation
merged_cellpose2 <- readRDS("/vast/projects/SpatialBench/segmentation/Cellpose/cellpose_updated_seurat/cellpose_seu_obj/dimension_reduction/SingleR_annotation/merged_cellpose2.rds")
```

# Visualsie merged data

## Cell type plot

```{r message=FALSE}
cell_plot_d <- DimPlot(merged_default, reduction = 'umap', group.by = 'cluster_label', label = TRUE) + 
  ggtitle("Cell types")
cell_plot_d
```

```{r message=FALSE}
cell_plot_b <- DimPlot(merged_baysor, reduction = 'umap', group.by = 'cluster_label', label = TRUE) + 
  ggtitle("Cell types")
cell_plot_b
```

```{r message=FALSE}
cell_plot_c <- DimPlot(merged_cellpose2, reduction = 'umap', group.by = 'cluster_label', label = TRUE) + 
  ggtitle("Cell types")
cell_plot_c
```

## Group plot

```{r message=FALSE}
grp_plot_d <- DimPlot(merged_default, reduction = 'umap', group.by = 'stim', alpha = 0.5) + 
  ggtitle("Groups")
grp_plot_d
```

```{r message=FALSE}
grp_plot_b <- DimPlot(merged_baysor, reduction = 'umap', group.by = 'stim', alpha = 0.5) + 
  ggtitle("Groups")
grp_plot_b
```

```{r message=FALSE}
grp_plot_c <- DimPlot(merged_cellpose2, reduction = 'umap', group.by = 'stim') + 
  ggtitle("Groups")
grp_plot_c
```


## Sample plot

```{r message=FALSE}
sample_plot_d <- DimPlot(merged_default, reduction = 'umap', group.by = 'sample_id') + 
  ggtitle("Samples")
sample_plot_d
```

```{r message=FALSE}
sample_plot_b <- DimPlot(merged_baysor, reduction = 'umap', group.by = 'sample_id') + 
  ggtitle("Samples")
sample_plot_b
```

```{r message=FALSE}
sample_plot_c <- DimPlot(merged_cellpose2, reduction = 'umap', group.by = 'sample_id') + 
  ggtitle("Samples")
sample_plot_c
```

## All-in-one

```{r fig.width=20, fig.height=5}
plots_d <- ggarrange(cell_plot_d, grp_plot_d, sample_plot_d, ncol = 3)
annotate_figure(plots_d, top = text_grob("Defaults segmentation", face = "bold", size = 20))
```

```{r fig.width=20, fig.height=5}
plots_b <- ggarrange(cell_plot_b, grp_plot_b, sample_plot_b, ncol = 3)
annotate_figure(plots_b, top = text_grob("Baysor segmentation", face = "bold", size = 20))
```

```{r fig.width=20, fig.height=5}
plots_c <- ggarrange(cell_plot_c, grp_plot_c, sample_plot_c, ncol = 3)
annotate_figure(plots_c, top = text_grob("Cellpose2 segmentation", face = "bold", size = 20))
```

# Pseudo-bulk analysis

## Aggregate counts to a sample level

```{r}
merged_default$samples <- paste0(merged_default$stim, merged_default$sample_id)
merged_baysor$samples <- paste0(merged_baysor$stim, merged_baysor$sample_id)
merged_cellpose2$samples <- paste0(merged_cellpose2$stim, merged_cellpose2$sample_id)

# NKs <- which(merged_cellpose$samples == "wt713" & merged_cellpose$manual_label == "NK")
# merged_cellpose@meta.data$manual_label[NKs] <- "Macrophages"
# saveRDS(merged_cellpose, "~/data/default_segmentation/merged_cellpose.rds")

# NKs <- which(merged_baysor$samples == "wt710" & merged_baysor$manual_label == "NK")
# merged_baysor@meta.data$manual_label[NKs] <- "Macrophages"
# saveRDS(merged_baysor, "~/data/baysor_segmentation/merged_baysor.rds")

# aggregate gene counts by cell types and groups
cts_d <- AggregateExpression(merged_default,
                             group.by = c("cluster_label", "samples"),
                             assays = 'SCT',
                             slot = "counts", # want the raw counts
                             return.seurat = FALSE)
cts_b <- AggregateExpression(merged_baysor,
                             group.by = c("cluster_label", "samples"),
                             assays = 'SCT',
                             slot = "counts", # want the raw counts
                             return.seurat = FALSE)
cts_c <- AggregateExpression(merged_cellpose2,
                             group.by = c("cluster_label", "samples"),
                             assays = 'SCT',
                             slot = "counts", # want the raw counts
                             return.seurat = FALSE)
```

## Data manipulation

```{r}
cts_d <- cts_d$SCT
cts_b <- cts_b$SCT
cts_c <- cts_c$SCT

# transpose
cts_d.t <- t(cts_d)
cts_b.t <- t(cts_b)
cts_c.t <- t(cts_c)

# convert to data.frame
cts_d.t <- as.data.frame(cts_d.t)
cts_b.t <- as.data.frame(cts_b.t)
cts_c.t <- as.data.frame(cts_c.t)

# get values where to split
splitRows <- gsub('_.*', '', rownames(cts_c.t))

# split data.frame - returns a list of separated data.frames
cts.split_d <- split.data.frame(cts_d.t,
                              f = factor(splitRows))
cts.split_b <- split.data.frame(cts_b.t,
                              f = factor(splitRows))
cts.split_c <- split.data.frame(cts_c.t,
                              f = factor(splitRows))

# fix colnames and transpose
cts.split.modified_d <- lapply(cts.split_d, function(x){
  rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x))
  t(x)
})
cts.split.modified_b <- lapply(cts.split_b, function(x){
  rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x))
  t(x)
})
cts.split.modified_c <- lapply(cts.split_c, function(x){
  rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x))
  t(x)
})
```

## Run DE analysis

```{r}
plot(density(counts_d[, 1]), ylim = c(0, 0.0006), main = "Raw counts of Cellpose samples")
for (i in 2:ncol(counts_d)) {
   den <- density(counts_d[, i])
   lines(den$x, den$y)
}
```

```{r}
plot(density(counts_b[, 1]), ylim = c(0, 0.0013), main = "Raw counts of Baysor samples")
for (i in 2:ncol(counts_b)) {
   den <- density(counts_b[, i])
   lines(den$x, den$y)
}
```

```{r}
plot(density(counts_c[, 1]), ylim = c(0, 0.0006), main = "Raw counts of Cellpose samples")
for (i in 2:ncol(counts_c)) {
   den <- density(counts_c[, i])
   lines(den$x, den$y)
}
```


### Functions 

```{r}
# perform CPM transformation on the data
transform_data <- function(cell_type) {
  # Get counts matrix
  counts_d <- cts.split.modified_d[[cell_type]]
  counts_b <- cts.split.modified_b[[cell_type]]
  counts_c <- cts.split.modified_c[[cell_type]]
  
  # generate sample level metadata
  colData <- data.frame(samples = colnames(counts_d))
  colData <- colData %>%
    mutate(condition = case_when(
      grepl("ctrl", samples) ~ "control",
      grepl("ko", samples) ~ "knockout",
      grepl("wt", samples) ~ "wildtype",
      TRUE ~ "unknown"  # in case none of the above match
    )) %>%
    column_to_rownames(var = "samples")
  
  # use limma to perform DE analysis
  # Create a DGEList with counts and sample group information
  dge_d <- DGEList(counts = counts_d, group = colData$condition)
  dge_b <- DGEList(counts = counts_b, group = colData$condition)
  dge_c <- DGEList(counts = counts_c, group = colData$condition)
  
  # CPM transformation to normalise different sequencing depths across samples
  dge_d$logcpm <- cpm(dge_d, log = TRUE)
  dge_b$logcpm <- cpm(dge_b, log = TRUE) 
  dge_c$logcpm <- cpm(dge_c, log = TRUE)
  
  return(list("default" = dge_d, "baysor" = dge_b, "cellpose2" = dge_c))
}

# density plot to check if samples share similar count distribution
den_plot <- function(data) {
  plot(density(data[, 1]), main = "", xlab = "log(CPM)", xlim = c(0, 22), ylim = c(0, 0.28))
  for (i in 2:ncol(data)) {
    den <- density(data[, i])
    lines(den$x, den$y)
  }
}

# Define marker genes for each cell type
marker_genes <- list(
  "B cells" = c("Cd22", "Bhlhe41", "Cd19", "Ighd", "Ighm", "Bcr", "Tbx21", 
                "Cxcr3", "Bbc3", "Bcl10", "Bcl2"),
  "T cells" = c("Cd4", "Cd3d", "Cd3e", "Cd8a", "Trac", "Foxp3", "Cxcr3", "Bbc3", 
                "Bcl10", "Bcl2"),
  "GC B cells" = c("Tbx21", "Pola1", "Nek2", "H2-Q6", "Tpx2", "Gpsm2", "Rrm1", 
                   "Ccnd3", "Akap12", "Tifa", "Cdca8", "Scn8a", "Pfn2", "Otub2", 
                   "Fcer2a", "Ifi30", "Stx11", "Cd40", "Ankrd33b", "Cd86", 
                   "Cd72", "Lmo4", "B3gnt5", "Serpinb6b", "Gstt2", "Actb", 
                   "Cd83", "Bcl6", "Aicda", "Rgs13", "Cxcr3", "Bbc3"),
  "Neutrophis" = c("Ngp", "S100a9", "Bbc3"),
  "Macrophages" = c("Cd274", "Cd68", "Cd209b", "Adgre1", "Csf1r", "Cd80", "Hmox1", "Bbc3"),
  "Plasma cells" = c("Cd38", "Jchain", "Cox6a2", "Siglech", "Bbc3"),
  "Erythrocytes" = c()
)

# Helper function: label false significant genes
mark_false_significant <- function(res, cell_type, marker_genes) {
  # Find markers for current cell type
  curr_markers <- marker_genes[[cell_type]]
  
  res <- res %>%
    mutate(Status = case_when(
      !(gene %in% curr_markers) & adj.P.Val < 0.05 & abs(logFC) > 1 ~ "False Significant",
      adj.P.Val < 0.05 & logFC < -1 ~ "Significant (Downregulated)",
      adj.P.Val < 0.05 & logFC > 1 ~ "Significant (Upregulated)",
      TRUE ~ "Not Significant"
    ))
  return(res)
}

de_analysis <- function(cell_type, groups = c("kovswt", "kovsctrl", "wtvsctrl")) {
  # Transform the aggregated count data
  trans_data <- transform_data(cell_type)
  dge_d <- trans_data$default
  dge_b <- trans_data$baysor
  dge_c <- trans_data$cellpose2
  
  # Filter lowly expressed genes
  keep_d <- filterByExpr(dge_d)
  keep_b <- filterByExpr(dge_b)
  keep_c <- filterByExpr(dge_c)
  dge_d <- dge_d[keep_d, , keep.lib.sizes = FALSE]
  dge_b <- dge_b[keep_b, , keep.lib.sizes = FALSE]
  dge_c <- dge_c[keep_c, , keep.lib.sizes = FALSE]
  
  # Calculate normalization factors (e.g., TMM normalization)
  dge_d <- calcNormFactors(dge_d, method = "TMM")
  dge_b <- calcNormFactors(dge_b, method = "TMM")
  dge_c <- calcNormFactors(dge_c, method = "TMM")
  
  # Construct design matrices
  design <- model.matrix(~0+group, data = dge_d$samples)
  colnames(design) <- gsub("group", "", colnames(design))
  
  # construct matrix of custom contrasts
  contr.matrix <- makeContrasts(
    kovswt = knockout - wildtype,
    kovsctrl = knockout - control,
    wtvsctrl = wildtype - control,
    levels = colnames(design))
  
  # Continue with voom and limma
  # transform count data to log2 counts-per-million (logCPM) to remove heteroscedascity
  v_d <- voom(dge_d, design)
  v_b <- voom(dge_b, design)
  v_c <- voom(dge_c, design)
  
  # fit linear model for each gene given a series of arrays for group comparison
  vfit_d <- lmFit(v_d, design)
  vfit_b <- lmFit(v_b, design)
  vfit_c <- lmFit(v_c, design)
  
  # compute contrasts from linear model fits to test specific hypothesis 
  vfit_d <- contrasts.fit(vfit_d, contrasts=contr.matrix)
  vfit_b <- contrasts.fit(vfit_b, contrasts=contr.matrix)
  vfit_c <- contrasts.fit(vfit_c, contrasts=contr.matrix)
  
  # apply empirical bayes shrinkage to standard errors for precise estimate of gene-wise variability
  efit_d <- eBayes(vfit_d)
  efit_b <- eBayes(vfit_b)
  efit_c <- eBayes(vfit_c)
  
  # view top genes from linear model fit
  # topTable(efit_c)
  # topTable(efit_b)
  
  # # Extract the decision results for the contrast of interest
  # testResults_c <- decideTests(efit_c)
  # testResults_b <- decideTests(efit_b)
  # 
  # # Identify the row indices for downregulated genes (where the test result is -1)
  # down_indices_c <- which(testResults_c[, groups] == -1)
  # down_indices_b <- which(testResults_b[, groups] == -1)
  # 
  # # Get the gene names corresponding to these indices
  # down_genes_c <- rownames(efit_c)[down_indices_c]
  # down_genes_b <- rownames(efit_b)[down_indices_b]
  # down_genes_c
  # down_genes_b
  
  # Extract the DE results for the contrast of interest
  res_d <- topTable(efit_d, coef = groups, number = Inf, adjust.method = "BH")
  res_b <- topTable(efit_b, coef = groups, number = Inf, adjust.method = "BH")
  res_c <- topTable(efit_c, coef = groups, number = Inf, adjust.method = "BH")
  res_d$gene <- rownames(res_d)
  res_b$gene <- rownames(res_b)
  res_c$gene <- rownames(res_c)
  
  # Label DE status including false significant genes
  res_d <- mark_false_significant(res_d, cell_type, marker_genes)
  res_b <- mark_false_significant(res_b, cell_type, marker_genes)
  res_c <- mark_false_significant(res_c, cell_type, marker_genes)
  
  return(list("default" = res_d, "baysor" = res_b, "cellpose2" = res_c))
}

de_res <- function(cell_type) {
  groups <- c("kovswt", "kovsctrl", "wtvsctrl")
  res <- lapply(groups, function(grp) de_analysis(cell_type, grp))
  res <- lapply(seq_along(groups), function(i) {
    res[[i]][["default"]]$segmentation <- "default"
    res[[i]][["baysor"]]$segmentation <- "baysor"
    res[[i]][["cellpose2"]]$segmentation <- "cellpose2"
    merged_res <- rbind(res[[i]][["default"]], res[[i]][["baysor"]], res[[i]][["cellpose2"]])
    merged_res$group <- groups[i]
    merged_res$group <- factor(merged_res$group, levels = groups)
    merged_res$segmentation <- factor(merged_res$segmentation, levels = c("default", "baysor", "cellpose2"))
    merged_res
  })
  do.call(rbind, res)
}

# visualise significant genes
volcano_plot <- function(res, cell_type) {
  ggplot(res, aes(x = logFC, y = -log10(adj.P.Val), color = Status)) +
    geom_point(size = 2, alpha = 0.8) +
    # Define colors: red for significant down, blue for significant up, gray for non-significant
    scale_color_manual(values = c("Significant (Downregulated)" = "orchid3",
                                "Significant (Upregulated)" = "skyblue3",
                                "False Significant" = "red",
                                "Not Significant" = "black")) +
    # Add reference lines: vertical line at 0 and horizontal line at p=0.05 threshold
    geom_vline(xintercept = -1, linetype = "dashed", color = "darkred") +
    geom_vline(xintercept = 1, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
    # Center the x-axis by setting limits from -3 to 3
    scale_x_continuous(limits = c(-4, 4)) +
    # Add plot title and axis labels 
    labs(title = cell_type,
         x = "Log2 Fold Change",
         y = "-Log10(adj P-value)",
         color = "DE Status") +
    facet_grid(rows = vars(group), cols = vars(segmentation)) + 
    # Use a theme with a border
    theme_bw() +
    theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
          plot.title = element_text(size = 16, hjust = 0.5, face = "bold"), # center the title
          strip.text = element_text(size = 12)) + 
    # Annotate the significant downregulated genes with their gene names
    geom_label_repel(
      data = subset(res, Status %in% c("Significant (Downregulated)", "Significant (Upregulated)")),
      aes(label = gene),
      box.padding = 0.35,
      point.padding = 0.5,
      point.size = 0.3,
      segment.color = "black",
      show.legend = FALSE, 
      size = 4)
}
```

### Density plots

Check the count distribution across samples.

According to the below density plots, the samples in all cell types have log(cpm) greater than 2. It seems like the dataset already filtered out lowly expressed genes. 

```{r}
cell_types <- unique(merged_default$cluster_label)
trans_data <- lapply(cell_types, function(cell_type) transform_data(cell_type))
names(trans_data) <- cell_types

for (cell_type in cell_types) {
  data <- trans_data[[cell_type]]
  den_plot(data$default$logcpm)
  title(main = paste("Transformed default data:", cell_type))
  den_plot(data$baysor$logcpm)
  title(main = paste("Transformed baysor data:", cell_type))
  den_plot(data$cellpose2$logcpm)
  title(main = paste("Transformed cellpose2 data:", cell_type))
}
```

### B cells

```{r fig.width=12, fig.height=8}
b_cell_res <- de_res("B cells")
volcano_plot(b_cell_res, "B cells")
```

### GC B cells

```{r fig.width=10, fig.height=8}
gc_b_cell_res <- de_res("GC B cells")
volcano_plot(gc_b_cell_res, "GC B cells")
```

### T cells

```{r fig.width=10, fig.height=8}
t_cell_res <- de_res("T cells")
volcano_plot(t_cell_res, "T cells")
```

### Macrophages

```{r fig.width=10, fig.height=8}
mac_res <- de_res("Macrophages")
volcano_plot(mac_res, "Macrophages")
```

### Neutrophils

```{r fig.width=10, fig.height=8}
neu_res <- de_res("Neutrophils")
volcano_plot(neu_res, "Neutrophils")
```

### Erythrocytes

```{r fig.width=10, fig.height=8}
erythrocyte_res <- de_res("Erythrocytes")
volcano_plot(erythrocyte_res, "Erythrocytes")
```

### Plasma cells

```{r fig.width=10, fig.height=8}
plas_cell_res <- de_res("Plasma cells")
volcano_plot(plas_cell_res, "Plasma cells")
```

## Volcano plots on selected cell types

Select B cells, GC B cells, and T cells on the kovswt comparison.

```{r fig.width=10, fig.height=8}
# Step 1: Define cell types and contrast
cell_types <- c("B cells", "GC B cells", "T cells")
contrast <- "kovswt"

# Step 2: Run DE analysis for each cell type (only kovswt)
de_list <- lapply(cell_types, function(ct) {
  res <- de_analysis(cell_type = ct, groups = contrast)
  
  # combine segmentation results and add metadata
  merged_res <- rbind(
    res[["default"]]  %>% mutate(segmentation = "Default"),
    res[["baysor"]]   %>% mutate(segmentation = "Baysor"),
    res[["cellpose2"]] %>% mutate(segmentation = "Cellpose2")
  )
  merged_res$cell_type <- ct
  merged_res
})

# Step 3: Combine all into one dataframe
res_all <- do.call(rbind, de_list)

# Ensure factors are ordered for consistent facet layout
res_all$segmentation <- factor(res_all$segmentation, levels = c("Default", "Baysor", "Cellpose2"))
res_all$cell_type <- factor(res_all$cell_type, levels = cell_types)

# Step 4: Plot all together
volcano_plots <- ggplot(res_all, aes(x = logFC, y = -log10(adj.P.Val), color = Status)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = c(
    "Significant (Downregulated)" = "orchid3",
    "Significant (Upregulated)" = "skyblue3",
    "False Significant" = "red",
    "Not Significant" = "black"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "darkred") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(
    title = "",
    x = "Log2 Fold Change",
    y = "-Log10(adj P-value)",
    color = "DE Status"
  ) +
  facet_grid(rows = vars(segmentation), cols = vars(cell_type)) +
  theme_bw() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 12),
    legend.position = "bottom"
  ) +
  geom_label_repel(
    data = subset(res_all, Status %in% c("Significant (Downregulated)", "Significant (Upregulated)")),
    aes(label = gene),
    box.padding = 0.35,
    point.padding = 0.5,
    point.size = 0.3,
    segment.color = "black",
    show.legend = FALSE,
    size = 3.5
  )
ggsave("figures/volcano_plots.pdf", plot = volcano_plots, width = 10, height = 8, dpi = 300)
volcano_plots
```

False significant genes

```{r fig.width=4, fig.height=3}
false_genes <- res_all %>% 
  filter(Status == "False Significant") %>%
  select(gene, segmentation, cell_type)

false_genes$segmentation <- factor(
  false_genes$segmentation,
  levels = c("Default", "Baysor", "Cellpose2"),
  labels = c("Cellpose1", "Baysor", "Cellpose2")
)

false_gene_counts <- false_genes %>% 
  group_by(segmentation, cell_type) %>% 
  summarise(counts = n(), .groups = "drop")

cols <- c("T cells" = "yellowgreen", 
          "Erythrocytes" = "violet", 
          "Neutrophils" = "mediumblue", 
          "B cells" = "tomato", 
          "Plasma cells" = "thistle1", 
          "GC B cells" = "yellow", 
          "Macrophages" = "skyblue2",
          "Endothelial cells" = "orange")

false_gene_count_plt <- ggplot(
  false_gene_counts, 
  aes(x = segmentation, y = counts, fill = cell_type)
) +
  geom_col() + 
  geom_text(
    aes(label = counts),
    # centers label within each bar segment
    position = position_stack(vjust = 0.5), 
    color = "black", 
    size = 3.5
  ) +
  scale_y_continuous(n.breaks = 8) + 
  scale_fill_manual(values = cols) + 
  labs(x = "Segmentations",
       y = "False significant gene counts",
       fill = "Cell types") + 
  theme_bw()
ggsave("figures/false_gene_count.pdf", plot = false_gene_count_plt,
       width = 4, height = 3, dpi = 300)
false_gene_count_plt
```

```{r fig.width=5, fig.height=5}
library(VennDiagram)
library(grid)

unique_false_genes <- false_genes %>% 
  group_by(segmentation) %>% 
  summarise(genes = list(unique(gene))) %>% 
  deframe()

# Aggregated False Significant Genes (B, GC B, T cells)
venn_false <- venn.diagram(
    x = unique_false_genes,
    category.names = c("Cellpose1", "Baysor", "Cellpose2"),
    filename = NULL,
    fill = scales::hue_pal()(3),
    alpha = 0.5,
    cex = 1.2,
    cat.cex = 1.2,
    main = ""
  )

pdf("figures/venn_false_genes.pdf", width = 5, height = 5)
grid.newpage()
grid.draw(do.call(grobTree, venn_false))
dev.off()
```

Find false significant genes commonly present in all segmentation methods

```{r}
Reduce(intersect, unique_false_genes)
```

Find genes only present in one segmentation method. 

```{r}
setdiff(
  unique_false_genes$Baysor,
  unlist(unique_false_genes[names(unique_false_genes) != "Baysor"])
)
```

```{r}
setdiff(
  unique_false_genes$Cellpose1,
  unlist(unique_false_genes[names(unique_false_genes) != "Cellpose1"])
)
```

```{r}
setdiff(
  unique_false_genes$Cellpose2,
  unlist(unique_false_genes[names(unique_false_genes) != "Cellpose2"])
)
```


## Venn Diagram

Aggregate false significant and significant genes across B cells, GC B cells and T cells. 

```{r fig.width=8, fig.height=4}
library(VennDiagram)
library(gridExtra)
library(dplyr)

aggregate_venn_across_celltypes <- function(cell_types = c("B cells", "GC B cells", "T cells"),
                                            contrast = "kovswt") {
  # Collect all DE results across cell types
  all_results <- lapply(cell_types, function(ct) {
    res_list <- de_analysis(ct, contrast)
    
    # Combine and add metadata
    merged_res <- rbind(
      res_list$default   %>% mutate(segmentation = "default",   cell_type = ct),
      res_list$baysor    %>% mutate(segmentation = "baysor",    cell_type = ct),
      res_list$cellpose2 %>% mutate(segmentation = "cellpose2", cell_type = ct)
    )
    merged_res
  }) %>% bind_rows()
  
  # --- Aggregate across cell types ---
  # True significant genes (union across cell types for each segmentation)
  sig_genes <- all_results %>%
    filter(adj.P.Val < 0.05, abs(logFC) > 1, Status != "False Significant") %>%
    group_by(segmentation) %>%
    summarise(genes = list(unique(gene))) %>%
    deframe()
  
  # False significant genes (union across cell types)
  false_genes <- all_results %>%
    filter(Status == "False Significant") %>%
    group_by(segmentation) %>%
    summarise(genes = list(unique(gene))) %>%
    deframe()
  
  # --- Create Venn diagrams ---
  venn_true <- venn.diagram(
    x = sig_genes,
    category.names = c("Default", "Baysor", "Cellpose2"),
    filename = NULL,
    fill = c("orchid3", "skyblue3", "gray70"),
    alpha = 0.5,
    cex = 1.2,
    cat.cex = 1.2,
    main = "Aggregated True Significant Genes (B, GC B, T cells)"
  )
  
  venn_false <- venn.diagram(
    x = false_genes,
    category.names = c("Default", "Baysor", "Cellpose2"),
    filename = NULL,
    fill = c("red", "orange", "darkred"),
    alpha = 0.5,
    cex = 1.2,
    cat.cex = 1.2,
    main = "Aggregated False Significant Genes (B, GC B, T cells)"
  )
  
  grid.arrange(grobTree(venn_true), grobTree(venn_false), ncol = 2)
}

# Run it
aggregate_venn_across_celltypes(c("B cells", "GC B cells", "T cells"), "kovswt") 
```


## UpSet plot

### Table of significant genes

Prepare data.frame for UpSet plot in KO vs WT comparison.

```{r}
signif_res <- list(b_cell_res, gc_b_cell_res, t_cell_res, mac_res, neu_res, erythrocyte_res, plas_cell_res)
cell_types <- c("B cells", "GC B cells", "T cells", "Macrophages", "Neutrophils", "Erythrocytes", "Plasma cells")
for (i in seq_along(signif_res)) {
  signif_ids <- grep("^Significant", signif_res[[i]]$Status)
  signif_res[[i]] <- signif_res[[i]][signif_ids, ]
  if (nrow(signif_res[[i]]) > 0) {
    signif_res[[i]]$celltype <- cell_types[i]
  }
}
signif_res <- do.call(rbind, signif_res)
rownames(signif_res) <- NULL

# sanity check
table(signif_res$Status)

# prepare table in wide format
signif_table <- signif_res %>% 
  filter(group == "kovswt") %>% 
  mutate(Regulation = case_when(
    Status == "Significant (Downregulated)" ~ "Down",
    Status == "Significant (Upregulated)" ~ "Up",
    TRUE ~ "other"
  )) %>% 
  select(gene, segmentation, celltype, Regulation) %>% 
  mutate(value = 1) %>% # marker for membership
  pivot_wider(
    # keep gene + regulation + segmentation as identifiers
    id_cols = c(gene, Regulation, segmentation),  
    names_from = celltype,          # make columns from cell types
    values_from = value,            # fill with 1s
    values_fill = list(value = 0)   # replace missing with 0
  )

# add any missing cell types which do not have DE genes
miss_id <- !cell_types %in% colnames(signif_table)[-c(1:3)]
miss_celltype <- cell_types[miss_id]
for (cell in miss_celltype) {
  signif_table[, cell] <- 0
}
```

### Plot

```{r}
default_signif <- signif_table %>% filter(segmentation == "default")
default_upset <- upset(default_signif, 
      intersect = colnames(default_signif)[-c(1:3)], # cell type columns
      sort_sets = FALSE,
      base_annotations = list(
        'Intersection size' = intersection_size(
          counts = TRUE,
          mapping = aes(fill = Regulation)
        ) + scale_fill_manual(values = c(
            "Down" = "orchid3", 
            "Up" = "skyblue3"
            ))
        ),
      width_ratio = 0.1) 
```

```{r}
baysor_signif <- signif_table %>% filter(segmentation == "baysor")
baysor_upset <- upset(baysor_signif, 
      intersect = colnames(baysor_signif)[-c(1:3)], # cell type columns
      sort_sets = FALSE,
      base_annotations = list(
        'Intersection size' = intersection_size(
          counts = TRUE,
          mapping = aes(fill = Regulation)
        ) + scale_fill_manual(values = c(
            "Down" = "orchid3", 
            "Up" = "skyblue3"
            ))
        ),
      width_ratio = 0.1) 
```

```{r}
cellpose2_signif <- signif_table %>% filter(segmentation == "cellpose2")
cellpose2_upset <- upset(cellpose2_signif, 
      intersect = colnames(cellpose2_signif)[-c(1:3)], # cell type columns
      sort_sets = FALSE,
      base_annotations = list(
        'Intersection size' = intersection_size(
          counts = TRUE,
          mapping = aes(fill = Regulation)
        ) + scale_fill_manual(values = c(
            "Down" = "orchid3", 
            "Up" = "skyblue3"
            ))
        ),
      width_ratio = 0.1) 
```

All-in-one

```{r fig.width=16, fig.height=5}
(default_upset | baysor_upset | cellpose2_upset) + plot_layout(guides = "collect")
```

Find downregulated genes in neutrophils.

```{r}
signif_table %>% 
  filter(Neutrophils == 1) %>% 
  select(gene, Regulation, segmentation, Neutrophils) %>% 
  group_by(gene, Regulation) %>% 
  summarise(count = n(), .groups = "drop") %>% 
  arrange(desc(count))
```

Find downregulated genes in Erythrocytes.

```{r}
signif_table %>% 
  filter(Erythrocytes == 1) %>% 
  select(gene, Regulation, segmentation, Erythrocytes) %>% 
  group_by(gene, Regulation) %>% 
  summarise(count = n())
```

B cells

Cell migration: Cxcr3

Macrophage markers: Adgre1, Cd209b

GC B cell marker: Akap12, Cd83

B cell marker: Bhlhe41

White pulp stromal cells: Ccl19

Endothelial cells: Egfl7

DC: Ppt1

```{r}
signif_table %>% 
  filter(`B cells` == 1) %>% 
  select(gene, Regulation, segmentation, `B cells`) %>% 
  group_by(gene, Regulation) %>% 
  summarise(count = n(), .groups = "drop") %>% 
  arrange(desc(count))
```

GC B cells

GC B cell markers: Akap12, Gstt2, Cd72, Fcer2a

Cell migration: Cxcr3

Macrophage markers: Adgre1

White blood cell/ everything: Bbc3

B cell marker: Ighd

```{r}
signif_table %>% 
  filter(`GC B cells` == 1) %>% 
  select(gene, Regulation, segmentation, `GC B cells`) %>% 
  group_by(gene, Regulation) %>% 
  summarise(count = n(), .groups = "drop") %>% 
  arrange(desc(count))
```


# Proprotion of B cells and GC B cells per sample

```{r}
cell_types_c <- merged_cellpose@meta.data[, c("manual_label", "nCount_Vizgen", "nFeature_Vizgen", "sample_id", "stim", "samples")]
cell_types_c$segmentation <- "cellpose"
cell_types_b <- merged_baysor@meta.data[, c("manual_label", "nCount_Vizgen", "nFeature_Vizgen", "sample_id", "stim", "samples")]
cell_types_b$segmentation <- "baysor"
cell_types <- rbind(cell_types_c, cell_types_b)

cell_prop <- cell_types %>% 
  group_by(segmentation, samples) %>% 
  summarise(
    b_cell_count = sum(manual_label == "B cells"), 
    gc_b_cell_count = sum(manual_label == "GC B cells"),
    total_cells = n(),
    b_cell_prop = b_cell_count / total_cells,
    gc_b_cell_prop = gc_b_cell_count / total_cells,
    .groups = "drop"
  )
```

GC B cell count across sample

```{r}
ggplot(cell_prop, aes(x = samples, y = gc_b_cell_count, fill = segmentation)) + 
  geom_col(position = "dodge") + 
  labs(x = "Samples",
       y = "GC B cell counts",
       fill = "Segmentation") + 
  theme_bw()
```

GC B cell proportion across sample

```{r}
ggplot(cell_prop, aes(x = samples, y = gc_b_cell_prop, fill = segmentation)) + 
  geom_col(position = "dodge") + 
  labs(x = "Samples",
       y = "Proportion of GC B cells",
       fill = "Segmentation") + 
  theme_bw()
```

Proportion of GC B cells per segmentation

```{r}
ggplot(cell_prop, aes(x = segmentation, y = gc_b_cell_prop, fill = segmentation)) +
  geom_boxplot() + 
  labs(x = "Segmentation",
       y = "Proportion of GC B cells") + 
  theme_bw() + 
  theme(legend.position = "none")
```

Counts per cell type

```{r fig.width=10, fig.height=10}
ggplot(cell_types, aes(x = samples, y = nCount_Vizgen, fill = segmentation)) +
  geom_violin(scale = "width", trim = FALSE) +
  geom_boxplot(aes(group = interaction(samples, segmentation)), 
               width = 0.2, outlier.shape = NA, fill = "white", color = "black",
               position = position_dodge(0.9)) +
  facet_wrap(~manual_label, ncol = 2, scales = "free_y") + 
  labs(x = "Samples",
       y = "Counts per cell",
       fill = "Segmentation") + 
  theme_bw()
```

Median counts of GC B cells per sample

```{r fig.width=3, fig.height=4}
counts_med <- cell_types %>% 
  filter(manual_label == "GC B cells") %>% 
  group_by(segmentation, samples) %>% 
  summarise(counts_med = median(nCount_Vizgen), .groups = "drop") %>% 
  ungroup()
ggplot(counts_med, aes(x = segmentation, y = counts_med, fill = segmentation)) +
  geom_boxplot() + 
  labs(x = "Segmentation",
       y = "Meidan counts of GC B cells per sample") + 
  theme_bw() +
  theme(legend.position = "none")
```

Genes per cell type

```{r fig.width=10, fig.height=10}
ggplot(cell_types, aes(x = samples, y = nFeature_Vizgen, fill = segmentation)) +
  geom_violin(scale = "width", trim = FALSE) +
  geom_boxplot(aes(group = interaction(samples, segmentation)), 
               width = 0.2, outlier.shape = NA, fill = "white", color = "black",
               position = position_dodge(0.9)) +
  facet_wrap(~manual_label, ncol = 2) + 
  labs(x = "Samples",
       y = "Counts per cell",
       fill = "Segmentation") + 
  theme_bw()
```

Median genes of GC B cells per sample

```{r fig.width=3, fig.height=4}
genes_med <- cell_types %>% 
  filter(manual_label == "GC B cells") %>% 
  group_by(segmentation, samples) %>% 
  summarise(genes_med = median(nFeature_Vizgen), .groups = "drop") %>% 
  ungroup()
ggplot(genes_med, aes(x = segmentation, y = genes_med, fill = segmentation)) +
  geom_boxplot() +
  labs(x = "Segmentation",
       y = "Median genes of GC B cells per sample") +
  theme_bw() +
  theme(legend.position = "none")
```



