---
title: "sub-clustering"
output: html_document
date: "2025-07-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/data/")
```

# Load packages

```{r}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(scater)
  library(scran)
  library(bluster)
  library(BiocParallel)
  library(dplyr)
  library(RColorBrewer)
})
```

# Load dataset

```{r}
ref <- readRDS("/stornext/General/data/user_managed/grpu_mritchie_1/Mei/SCORE/ST_MD/analysis/output/RDS/sce.labelled.RDS")
```

# Sub-clustering B cell clusters

Extract B cell clusters.

```{r}
b_cells <- ref[, ref$cluster_label == "B cells"]
```

Normalise and identify variable genes as the number of B cells is large.

```{r}
# model gene variance
mgv <- modelGeneVar(b_cells)
b_cells <- denoisePCA(b_cells, technical = mgv, subset.row = getTopHVGs(mgv, n = 7000))
```

Dimensionality reduction and clustering.

```{r}
# UMPA
b_cells <- runUMAP(b_cells, dimred = "PCA")
```

Use Leiden clustering method with resolution option.

```{r}
# set up parallelism
param <- MulticoreParam(workers = 4)
```


```{r}
# set up k and resolution values for parameter tuning
ks <- seq(30, 70, 10)
res <- seq(0.3, 1, 0.1)
results <- list()

for (k in ks) {
  for (r in res) {
    cluster <- clusterRows(
      reducedDim(b_cells, "PCA"), 
      BLUSPARAM = SNNGraphParam(k = k, cluster.fun = "leiden", 
                                cluster.args = list(resolution = r), 
                                BPPARAM = param)
    )
    key <- paste0("k", k, "_res", r)
    results[[key]] <- cluster
  }
}
```

```{r}
# plot clusters on UMAP
for (key in names(results)) {
  b_cells$subcluster <- results[[key]]
  print(plotUMAP(b_cells, colour_by = "subcluster") + ggtitle(key))
}
```


```{r}
b_cells$subcluster <- results[["k60_res0.6"]]
plotUMAP(b_cells, colour_by = "subcluster") + ggtitle("k60_res0.6")
```

```{r}
markers <- findMarkers(b_cells, groups = b_cells$subcluster)

top_markers <- lapply(
  markers, 
  function(cluster) {
    as.data.frame(cluster) %>% 
      filter(FDR < 0.05 & abs(summary.logFC > 1)) %>% 
      arrange(desc(abs(summary.logFC))) %>% 
      slice_head(n = 20)
  }
)
```

```{r}
b_cell_markers <- c("Cd22", "Bhlhe41", "Cd19", "Ighd", "Ighm")

gcb_cell_markers <- c("Tbx21", "Pola1", "Nek2", "H2-Q6", "Tpx2", "Gpsm2", "Rrm1", 
                      "Ccnd3","Akap12", "Tifa", "Cdca8", "Scn8a", "Pfn2", 
                      "Otub2", "Fcer2a", "Ifi30", "Stx11", "Cd40", "Ankrd33b", 
                      "Cd86", "Cd72", "Lmo4", "B3gnt5", "Serpinb6b", "Gstt2", 
                      "Actb", "Cd83", "Bcl6", "Aicda", "Rgs13")

plasma_cell_markers <- c("Cd38", "Jchain")
```

```{r}
plotExpression(b_cells, features = b_cell_markers, 
               x = "subcluster",
               colour_by = "subcluster") 
```

Cluster 1: Pola1, Aicda, Rgs13

```{r}
for (i in seq(1, length(gcb_cell_markers), 6)) {
  print(plotExpression(b_cells, features = gcb_cell_markers[i:(i+5)], 
               x = "subcluster",
               colour_by = "subcluster") )
}
```
```{r}
plotReducedDim(b_cells, dimred = "UMAP", colour_by = "Pola1")
```

```{r}
plotReducedDim(b_cells, dimred = "UMAP", colour_by = "Aicda")
```

```{r}
plotReducedDim(b_cells, dimred = "UMAP", colour_by = "Rgs13")
```

Cluster 6: Jchain

```{r}
plotExpression(b_cells, features = plasma_cell_markers, 
               x = "subcluster",
               colour_by = "subcluster") 
```

```{r}
plotReducedDim(b_cells, dimred = "UMAP", colour_by = "Jchain")
```

# Assign refined cell labels back to original data

Assign cell labels to sub-cluster

```{r}
b_cells$subcluster_labels <- "B cells"
b_cells$subcluster_labels[b_cells$subcluster == 1] <-  "GC B cells"
b_cells$subcluster_labels[b_cells$subcluster == 6] <-  "Plasma cells"
b_cells$subcluster_labels <- as.factor(b_cells$subcluster_labels)
```

```{r}
plotReducedDim(b_cells, dimred = "UMAP", colour_by = "subcluster_labels")
```

Assign defined GC and plasma cell labels back to original data.

```{r}
sub_labels <- b_cells$subcluster_labels
names(sub_labels) <- colnames(b_cells)

ref$fine_labels <- as.character(ref$cluster_label)
colData(ref)[names(sub_labels), "fine_labels"] <- as.character(sub_labels)
ref$fine_labels <- as.factor(ref$fine_labels)
```

Check on dimension reduction plot.

```{r}
palette <- brewer.pal(12, "Paired")
plotReducedDim(ref, dimred = "UMAP", color_by = "fine_labels") + 
  scale_color_manual(values = palette)
```

# Granulocytes & neutrophils

Neutrophil markers: Ngp, S100a9

Granulocyte cluster has high expression level of neutrophil markers. 

```{r}
plotReducedDim(ref, dimred = "UMAP", colour_by = "Ngp")
plotReducedDim(ref, dimred = "UMAP", colour_by = "S100a9")
```

```{r}
plotExpression(ref, features = c("Ngp", "S100a9"), 
               x = "fine_labels", colour_by = "fine_labels") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```


# Stem cells & endothelial cells

Endothelial cell markers: Egfl7, Madcam1

```{r}
plotReducedDim(ref, dimred = "UMAP", colour_by = "Egfl7")
plotReducedDim(ref, dimred = "UMAP", colour_by = "Madcam1")
```
Check expression level of nuetrophil marker Egfl7 in stem cell cluster.

```{r}
plotExpression(ref, features = "Egfl7", 
               x = "fine_labels", colour_by = "fine_labels") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none")
# cannot map stem cell cluster to endothelial cells
```

# Save refined cell labels.

```{r}
saveRDS(ref, file = "sce.labelled.fine.RDS")
```

